Non-cognate immunity proteins provide broader defenses against interbacterial effectors in microbial communities

Version 1

Abstract

Abstract

Dense microbial communities, like the gut and soil microbiomes, are dynamic societies. Bacteria can navigate these environments by deploying proteins (effectors) that alter foreign cells’ behavior. Immunity proteins preferentially protect neighboring sibling cells, in contrast to canonical toxin-antitoxin systems. A prevailing hypothesis is that when immunity proteins are bound to specific (cognate) protein partners, it is sufficient to disrupt their function; further, there is little-to-no crosstalk with other (non-cognate) effectors. Here, we build on sporadic reports challenging these hypotheses. We show that immunity proteins from a newly defined protein family can bind and protect against non-cognate PD-(D/E)XK-containing effectors from diverse phyla. We describe the domains essential for binding and function and show that binding alone is insufficient for protective activity. Moreover, we found that these effector and immunity genes co-occur in individual human microbiomes. These results expand the growing repertoire of bacterial protection mechanisms and force us to reconsider how non-cognate interactions impact community structure within complex ecosystems.

Introduction

Specificity between protein-protein interactions is key for several biological processes, such as metabolism, development, and intercellular signaling. Binding to an incorrect partner or disrupted binding of the specific (cognate) protein can cause a diseased state or cell death1. For bacterial social behaviors, cognate protein-protein interactions between cells impact organism fitness and community structure, such as excluding foreign cells2. The importance of specificity between cognate partners in social behaviors remains largely unchallenged. In other contexts, flexible (“promiscuous”) binding allows protein partners to retain their interactions when undergoing rapid mutational changes, such as during immune recognition of viral particles3, 4. Unknown is whether flexible binding between noncognate proteins can occur during bacterial social behaviors. An expanded protective function would reveal new bacterial behaviors that influence individual fitness and community structure in dynamic ecosystems.

Microbes often exist within dense, multi-phyla communities, like the human gut microbiome, where they communicate and compete with neighbors. Bacteria can use effector-immunity protein (EI) pairs in these environments to gain advantages5, 6. Unlike toxin-antitoxin systems in which a single cell produces both toxic and neutralizing proteins, bacteria inject cell-modifying proteins (effectors) directly into neighboring cells via several contact-dependent transport mechanisms, including the type VI secretion system (T6SS), type IV secretion system (T4SS), and contact-dependent inhibition (CDI)7, 8. Clonal siblings produce the matching immunity protein that modifies the effector’s activity. For lethal effectors, both clonal and non-clonal cells are negatively impacted when binding is disrupted or absent9. These interactions between EI pairs can shape community composition by changing bacterial fitness.

The binding specificity between matching EI pairs has historically defined immunity protein protection, but recent studies raise doubts. Antitoxins have multiple mechanisms for neutralizing their cognate toxin, usually within a single cell. By contrast, EI pairs interact within a neighboring cell which puts unique restrictions on both their protection mechanisms and their evolution10. The predominant model is that T6SS-associated EI pairs act like a tumbler lock-and-key, where each effector protein has a single cognate partner11. Most immunity proteins bind their cognate effectors, often at the active site, to neutralize effector activity12. As such, EI pairs typically have increased protein-protein specificity between matching pairs to protect sibling cells13. However, experiments with engineered proteins reveal that changes to an immunity protein could allow it to bind effectors other than its cognate partner14, 15. Also, the T6SS-associated effector and immunity proteins from Salmonella enterica subsp. enterica serovar Typhimurium and Enterobacter cloacae, which are phylogenetically close, bind each other in vitro and protect against the other in vivo16. These studies suggest that the predominant model needs to be revised to explain the breadth of immunity protein protection.

We studied an EI pair in Proteus mirabilis to examine the prevailing model. This opportunistic pathogen resides in human and animal guts and can cause recurrent and persistent urinary tract infections17. P. mirabilis encodes a lethal T6SS-dependent EI pair that impacts competition and collective motility18. The molecular functions of the putative effector and immunity proteins remained unknown. Here, we characterized this EI pair and determined the critical residues for activity, leading to the discovery of two protein families. We showed that these immunity proteins bind non-cognate effectors produced by bacteria from different phyla. Our results indicated that binding was insufficient to neutralize effector proteins within this immunity protein family. Further, we found that these flexible EI pairs from various phyla naturally co-occur in individual human microbiomes. This work provides compelling evidence for cross-protection and supports a fundamental reinterpretation of EI pairs and their ecological significance.

Results

RdnE is a DNA nuclease and seeds a PD-(D/E)XK subfamily

To compete against other strains, P. mirabilis strain BB2000 needs both the idrD gene and the T6SS, suggesting that the idrD-encoded protein functions as a T6SS-associated effector18. Similar effectors often contain an enzymatic domain in the C-terminus19, 20. As a result, we investigated the function of the final 138 amino acids at IdrD’s C-terminus, now named RdnE for recognition DNA nuclease effector. We measured bacterial growth using a strain derived from BB2000 that has disruptions in its native idrD and downstream genes18. These P. mirabilis cells had 1000 fewer cells per mL when engineered to overproduce RdnE in trans than the negative control containing no RdnE protein (Figure 1A). An equivalent growth pattern occurred in Escherichia coli cells under the same conditions (Supplemental Figure 1). Thus, RdnE was lethal in vivo. RdnE’s initial 86 amino acids contain a PD-(D/E)XK motif, which is suggestive of nucleotide degradation. The PD-(D/E)XK superfamily includes proteins with broad functions, including effectors that degrade DNA or RNA21–23. Three residues in the catalytic site—D, D/E, and K—are required for activity24. Therefore, we changed the corresponding residues in RdnE (D39, E53, and K55) to alanine, separately and together.

RdnE homologs act as DNA endonucleases and contain interchangeable domains.

A) Cell viability (colony forming units per mL) after protein production in swarms of P. mirabilis strain idrD*, which does not produce RdnE and RdnI. Cells produced GFPmut2, RdnE, or mutant variants in the predicted PD-(D/E)XK motif: D39A, E53A, K55A, or all. B) In vitro DNA degradation assay for ProteusRdnE. Increasing concentrations of a negative control, ProteusRdnE-FLAG, or ProteusRdnED39A-FLAG were incubated with methylated or unmethylated lambda DNA (48,502 bp) and analyzed by gel electrophoresis. Plasmid DNA degradation is in Supplemental Figure 1. C) In vitro DNA degradation assay for domain deletions of ProteusRdnE. The first construct removed the first alpha helix without disturbing the catalytic residues, and the second construct contained the PD-(D/E)XK motif and removed region 2. Increasing concentrations were analyzed as in (B). D) Multiple sequence alignment between P. mirabilis and R. dentocariosa RdnE sequences. Black bar show the PD-(D/E)XK motif and the grey bar marks the variable region 2 domain. Conserved residues are highlighted in dark blue. Secondary structure predictions identified using Ali2D55, 56 (h for alpha helix, e for beta sheet); the catalytic residues (stars) are noted above the alignment. (E,F) In vitro DNA degradation assay and analysis as in (B). (E) Increasing concentrations of either a negative control, RothiaRdnE-FLAG, or RothiaRdnED39A-FLAG. (F) The PD-(D/E)XK motifs were swapped between the RothiaRdnE (orange) and the ProteusRdnE (green) sequences and compared to the wild-type RdnE proteins.

P. mirabilis producing these mutant proteins showed growth equivalent to the negative control lacking RdnE (Figure 1A). We also noticed that E. coli cells that were producing RdnE had morphologies that were indicative of DNA damage or stress. These cells were elongated, and the DAPI-stained DNA was distributed irregularly within the cells (Supplemental Figure 1). Cells producing a defunct D39A mutant (RdnED39A) did not have this appearance (Supplemental Figure 1). Therefore, the PD-(D/E)XK motif was essential for growth arrest.

The importance of the PD-(D/E)XK motif for activity suggested that RdnE was a nuclease, but defining its molecular target required direct analysis. We synthesized RdnE with a C-terminus FLAG epitope tag using in vitro translation because of its lethality (Supplemental Figure 1). We measured the protein concentrations, added phage lambda DNA (methylated or unmethylated) to progressively higher RdnE protein concentrations, and then performed agarose gel electrophoresis analysis. Degradation of lambda DNA occurred in the presence of RdnE, regardless of the DNA methylation state (Figure 1B). The RdnED39A construct caused a slight reduction in lambda DNA, while the negative control showed no DNA loss (Figure 1B). RdnE also caused a reduction in plasmid DNA, indicating its endonuclease activity (Supplemental Figure 1). These results revealed that RdnE caused DNA degradation in vitro in a PD-(D/E)XK-dependent manner.

RdnE likely has two different domains, as there is a region directly following the PD-(D/E)XK motif. A two-domain architecture is similar to that described for DNases25, 26 and other T6SS effectors like some PoNe-containing DNases21. We made independent deletions of each potential RdnE domain to assess whether both were essential for DNase activity. One construct deleted the first alpha helix without disturbing the catalytic residues; the other deleted the region after the PD-(D/E)XK motif, which we termed “region 2.” The resulting proteins, produced via in vitro translation, were assayed for DNase activity as described above. The truncated proteins resulted in no loss of lambda DNA (Figure 1C), indicating that both domains are necessary for degradation activity.

We next asked whether RdnE homologs also act as DNA nucleases. A bioinformatics search revealed the closest RdnE homolog (outside of Proteus) was found in the Actinobacteria, Rothia dentocariosa C6B. Rothia species are inhabitants of the normal oral flora, dwelling in biofilms within the human oral cavity and pharynx27. The two RdnE proteins (ProteusRdnE and RothiaRdnE) share approximately 55% amino acid sequence identity, mostly within the PD-(D/E)XK domain, and have similar predicted secondary structures (Figure 1D). Given this, we hypothesized that RothiaRdnE also acted as a DNA nuclease.

We analyzed RothiaRdnE for PD-(D/E)XK-dependent DNA nuclease activity. We produced it and a predicted null mutant, RothiaRdnED39A, using in vitro translation. Samples containing the RothiaRdnED39A protein or a negative control had similar DNA levels (Figure 1E). By contrast, samples with the wild-type RothiaRdnE protein showed a loss of lambda DNA regardless of methylation state, indicating that RothiaRdnE also had DNA nuclease activity (Figure 1E). To further analyze RdnE’s architecture, we exchanged region 2 between the ProteusRdnE and RothiaRdnE sequences and assayed for nuclease activity. The hybrid proteins degraded lambda DNA, unlike the negative control (Figure 1F). While DNA degradation by these trans-phyla hybrid proteins required more protein than the wild-type enzymes, the cross-phyla protein fragments could complement one another. These data indicated that these RdnE proteins form a novel PD-(D/E)XK-containing DNA nuclease subfamily.

RdnI binds and neutralizes RdnE

Effectors have cognate immunity proteins, often adjacently located on the chromosome. We hypothesized that the gene adjacent to rdnE in BB2000, rdnI (previously named “idrE”), encodes the cognate immunity protein (Figure 2A). RdnI did not have defined domains, and its function was unknown. We assessed RdnI’s activity using microscopic and cell growth analysis. Swarming P. mirabilis cells are normally elongated with DAPI-stained DNA found along the cell body (Figure 2B). By contrast, swarming cells producing RdnE in trans did not elongate, had a reduced DAPI signal, and had an accumulation of misshapen cells (Figure 2B). Cell shape and DNA-associated fluorescence levels returned to normal when cells concurrently produced the RdnE and RdnI proteins (Figure 2B). RdnI production also rescued cell growth in E. coli cells producing RdnE (Supplemental Figure 2). These data suggest that RdnI inhibits RdnE’s lethality.

RdnI binds to and protects against RdnE in vivo and in vitro.

A) Domain architecture for the idr locus in P. mirabilis strain BB200018. At the top are genes with Pfam domains listed below them. Gray boxes denote PAAR and Rhs domains in the N-terminal region of the full-length IdrD protein. B) Micrographs of P. mirabilis strain idrD* cells carrying an empty vector, a vector for producing RdnE, or a vector for producing RdnE and RdnI. DNA was visualized by DAPI stain. Phase, left; fluorescence, right. C) Swarm competition assay18 of wild-type P. mirabilis strain BB2000 (donor) competed against the vulnerable target, which is P. mirabilis strain ATCC29906 carrying an empty vector, a vector for producing RdnI-StrepII, or a vector for producing GFP. Left, schematic of swarm competition assay where top left colony is BB2000, top right colony is ATCC29906 with its vector cargo, and bottom colony is a 1:1 mixture of BB2000 and ATCC29906 with its vector cargo. Gray boxes underneath indicate whether BB2000 (top) or ATCC29906 (bottom) dominated in the 1:1 mixture and white arrows point to the boundary line that forms between different strains. D) Bacterial two-hybrid (BACTH) assay with RdnED39A-FLAG, RdnI-StrepII, and GFPmut2. The colorimetric change was discerned in the presence of the substrate Xgal and inducer IPTG. E) An anti-FLAG batch co-immunoprecipitation of RdnED39A-FLAG and RdnI-StrepII. RdnED39A-FLAG or exogenous FLAG-BAP (soluble fraction) was incubated with anti-FLAG resin (FLAG flow through). RdnI-StrepII was then added to the resin (RdnI-StrepII flow through). Any proteins bound to resin were eluted with FLAG-peptide (Elution) and analyzed by anti-FLAG and anti-StrepII western blots.

To determine whether RdnI provided protection against injected RdnE within mixed communities, we used swarm competition assays, which combine one-to-one mixtures of P. mirabilis strains and then measure dominance18. The control strain was wild-type strain BB2000 (herein called “BB2000”), which naturally produced RdnE and RdnI. The other was strain ATCC29906, which did not naturally produce RdnE and RdnI. These two strains formed a visible boundary between swarming monoculture colonies (Figure 2C). The mixed-strain colony merged with BB2000 in one-to-one competitions, demonstrating BB2000’s dominance (Figure 2C). A similar outcome was seen when ATCC29906 was producing Green Fluorescent Protein (GFPmut2). However, BB2000 did not outcompete ATCC29906 engineered to produce RdnI with a C-terminal StrepII epitope tag (“RdnI-StrepII”), seen by a merging of the mixed-strain colony with ATCC29906 (Figure 2C). Thus, RdnI protected cells against injected RdnE within mixed communities.

Based on the prevailing model, we predicted that a cognate EI pair should bind to one another, which we evaluated in vivo and in vitro. We used the defunct mutant (RdnED39A-FLAG) for these assays because producing the wild-type RdnE protein kills cells. For in vivo analysis, we used bacterial two-hybrid assays (BACTH) in which the reconstitution of the T18 and T25 fragments of adenylate cyclase results in the colorimetric change to blue in the presence of the substrate, X-gal28, 29. Constructed vectors contained genes for RdnED39A-FLAG, RdnI-StrepII, or GFPmut2 on the C-termini of the T18 or the T25 fragments. When the reporter strain produced RdnED39A-FLAG or RdnI-StrepII with GFPmut2, the resultant yellow color was equivalent to when X-gal was absent (Figure 2D). There was also minimal color change when an individual protein was produced on both fragments (Figure 2D). However, the reporter strains made blue colonies when X-gal was present and the cells concurrently produced RdnED39A-FLAG and RdnI-StrepII. These results indicated that RdnE and RdnI bind to each other in vivo. We used batch in vitro co-immunoprecipitation assays to confirm the in vivo binding result. Separate E. coli strains produced either RdnED39A-FLAG or had a negative control, exogenous FLAG-BAP (E. coli bacterial alkaline phosphatase with a FLAG epitope tag) added to cell lysate. An anti-FLAG western blot showed both FLAG-BAP (∼ 50 kDa) and RdnED39A-FLAG (∼ 17 kDa) in the soluble and elution fractions. RdnI-StrepII eluted with RdnED39A-FLAG but not the negative control (Figure 2E). The western blot results corresponded with the Coomassie blue-stained gels (Supplemental Figure 2). Overall, our data showed that Proteus RdnE and RdnI form a cognate EI pair. However, questions about their prevalence among bacteria remained, because RdnE and RdnI matched no known effector or immunity families.

Expansion of the RdnE and RdnI protein families revealed similar gene architecture and secondary structures

Gene neighborhood analysis can guide homology inference and protein comparisons. We conducted consecutive searches with BLAST30 and HMMER31 to identify sequences that encoded proteins with high similarity to RdnE and RdnI (Supplemental Figure 3). The final list contained 21 EI pairs from a variety of phyla (Table 3). Although the genes surrounding these putative EI pairs differed, many shared mobile and transport-associated elements, such as recombinant hotspot (Rhs) sequences or other similar peptide-repeat sequences (Figure 3A). Several gene neighborhoods had transport-associated genes, such as the T6SS-associated vgrG/tssC gene and the CDI-associated cdiB gene. A few also included putative immunity proteins from other reported families, like immunity protein 44 (Pfam15571) in Taylorella asinigenitalis MCE3 and immunity protein 51 (Pfam15595) in Chryseobacterium populi CF314. Notably, these organisms varied widely in origin and residence. Some were from the soil rhizosphere (Pseudomonas ogarae and C. populi) and others from the human microbiome (P. mirabilis, R. dentocariosa, and Prevotella jejuni) (Figure 3A). The phylogenetic tree showed distribution across bacterial phyla (Figure 3B), perhaps reflecting these proteins’ role in community interactions.

RdnE and RdnI protein families share conserved residues and predicted structures.

A) Gene neighborhoods for RdnE and RdnI homologs. Listed are gene neighborhoods, relevance, and niche, which we identified using IMG/M from the Joint Genomics Institute. Colors highlight conserved function/genes (not to scale). (Agr: Agriculture, Med: Medical, Env: Environmental), and the site of isolation. B) Phylogenetic tree based on NCBI taxonomy. Scale is located below the graph. The colored circles represent phyla (green: Actinobacteriota; yellow: Firmicutes; blue: Bacteroidota; pink: Proteobacteria). C) Unrooted maximum likelihood trees of the RdnE (left) and RdnI (right) homologs. Trees were created with RaxML60, and the scale is annotated below. The colored circles represent phyla (same as in B). D) Protein alignments overlaid with either predicted secondary structures (top) or conserved residues (bottom) of the RdnE and RdnI homologs. MUSCLE alignments52 are highlighted by secondary structures (red: alpha helices, light blue: beta sheets), or conserved residues (dark blue). White represents gaps in the protein alignment. The bars below mark the predicted conserved and variable domains. E) Alignments of AlphaFold2 predictions for RdnE and RdnI sequences from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), and P. ogarae (dark blue). Structures were generated using ColabFold32 and aligned using PyMol. The P. mirabilis RdnI sequence is a natural variant with two residue substitutions (V101E and I222M) as compared to BB2000 (supplemental figure 6).

We next constructed maximum likelihood trees to examine the relationship between the identified proteins. Both the RdnE and RdnI trees diverged from the species tree (Figure 3B) and each other (Figure 3C). Proteins from evolutionarily distant bacteria appeared to share more similarities compared to those from more closely related genera. These results are consistent with potential horizontal gene transfer seen for other EI pairs7.

Yet, despite the differences in phyla and gene neighborhoods, the RdnE and RdnI proteins displayed the defining characteristics of protein families. The predicted secondary structures of the RdnE proteins were similar even though the amino acid sequences differed (Figure 3D, Supplemental Figure 3). The RdnE proteins showed two distinct domains, as with the Proteus and Rothia results (Figure 1): the PD-(D/E)XK region followed by a sequence variable region (region 2). Further, the predicted tertiary structures were consistent with PD-(D/E)XK folds (three β-sheets flanked by two α-helices, α/β/α) found in other proteins (Figure 3E, Supplemental Figure 4)24, 32. These findings suggested that domains in RdnE are conserved across diverse phyla and the sequences seed a distinct PD-(D/E)XK subfamily.

While immunity proteins within a family have diverse overall amino acid sequences, conserved secondary structures and some conserved residues are not uncommon among immunity protein families. Indeed, they are often used to characterize these families33. We found that, while the RdnI proteins shared minimal primary amino acid sequence identity, they were predicted to contain several α-helices (Figure 3D, Supplemental Figure 3) and had similar AlphaFold2-predicted tertiary structures (Figure 3E, Supplemental Figure 4). We also discovered a region with three alpha-helices and several conserved residues, which we named the “conserved motif” (Figure 3D). Interestingly, the addition of this conserved motif indicates that the RdnI architecture mirrors that of the two-domain RdnE proteins (Figure 3D). Therefore, the RdnI conserved motif may also indicate a novel immunity domain and seed a protein family.

Binding flexibility in RdnI allows for cross-species protection

We deployed a structure-function approach to determine the conserved motif’s role in the ProteusRdnI’s activity. Bioinformatics using AlphaFold232 and Consurf34 revealed seven highly conserved residues within this region; four of these (Y197, H221, P244, E246) clustered together within the AlphaFold2 structure and are identical between sequences (Figures 4A). In a sequence-optimized (SO) RdnI, we individually changed each of these four to alanine and discovered that each alanine-substituted variant behaved like the wildtype and inhibited RdnE activity (Supplemental Figure 5). We then replaced all seven residues (Y197, S235, K258, and the original four) with alanine (ProteusRdnI7mut-StrepII) and found that, unlike wild-type ProteusRdnI-StrepII, this construct was not protective in swarm competition assays (Figure 4B). However, the ProteusRdnI7mut-StrepII mutant still bound RdnED39A-FLAG in bacterial two-hybrid assays (Figure 4C). Therefore, the seven residues in the conserved motif are critical for RdnI’s neutralizing function but dispensable for binding RdnE.

The RdnI protein family can offer cross-protection due to an interchangeable conserved domain that is critical for function.

A) Sequence logo of the RdnI’s conserved motif. Stars indicate the seven analyzed residues. B) Swarm competition assay with ATCC29906 producing either RdnI-StrepII or RdnI7mut-StrepII, which contains mutations in all seven conserved residues. We used a sequence-optimized (SO) RdnI protein that had a higher GC% content and an identical amino acid sequence for ease of cloning. Left: schematic of swarm competition assay as in Figure 2. Gray boxes indicate which strain dominated over the other. White arrows point to the boundary formed between different strains. C) BACTH assay of RdnED39A-FLAG with SO RdnI-StrepII or RdnI7mut-StrepII. GFPmut2 was used as a negative control. D) Swarm competition assay with ATCC29906 expressing either the wild-type RdnI or a RdnI truncation. The three truncations were in the first alpha helix (amino acids 1-85), the second half of RdnI (amino acids 150-305), and the end of the protein (amino acids 235-305). E) BACTH assay of RdnED39A-FLAG with wild-type RdnI and the three RdnI truncations. F) Swarm competition assay with ATCC29906 expressing foreign RdnI proteins. G) BACTH assay of RdnED39A-FLAG with each of the foreign RdnI proteins. GFPmut2 was used as a negative control. H) Swarm competition assay with ATCC29906 producing SO RdnI with swapped conserved motifs. I) BACTH assay of RdnED39A-FLAG with SO RdnI with swapped conserved motifs. Colored bars denote RdnI-StrepII proteins from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), or P. ogarae (dark blue).

Given that the conserved motif and nearby regions are likely involved in function, we queried for potential functions in the remainder of the RdnI protein. We engineered variants that were either (1) the first 85 amino acids, (2) amino acids 150 to 305, which contained an intact conserved motif, or (3) amino acids 235 to 305, which contained the last alpha helix of the conserved motif (Supplemental Figure 3). None of these constructs protected against RdnE’s lethality in vivo (Figure 4D), demonstrating that the entire protein is likely essential for function. The variant containing the first 85 amino acids of ProteusRdnI was the only construct to bind ProteusRdnE, indicating that the N-terminal region is sufficient for binding between this P. mirabilis EI pair (Figure 4E). Thus, binding is necessary but insufficient for neutralization, and the neutralization activity likely resides within the second half of RdnI. These results, as well as the functional domain swaps between the Proteus and Rothia RdnE proteins (Figure 1F), contradicted the prevalent model’s assertion that one-to-one binding is both necessary and sufficient for immunity protein function against a cognate effector.

Therefore, we explored the relationship between non-cognate RdnE and RdnI proteins from various phyla. We first asked whether non-cognate RdnI immunity proteins could protect against injected ProteusRdnE. Using the swarm competition assays, we competed BB2000 against ATCC29906 engineered to produce RdnI homologs from R. dentocariosa, P. jejuni, or P. ogarae (RothiaRdnI-StrepII, PrevotellaRdnI-StrepII, and PseudomonasRdnI-StrepII, respectively). BB2000 dominated the swarm when ATCC29906 produced GFPmut2, PrevotellaRdnI-StrepII, or PseudomonasRdnI-StrepII (Figure 4F). However, ATCC29906 outcompeted BB2000 when making ProteusRdnI-StrepII or RothiaRdnI-StrepII (Figure 4F). Consistent with these findings, the RdnI immunity proteins from Proteus and Rothia consistently bound ProteusRdnED39A in in vivo and in vitro assays, but the Prevotella and Pseudomonas variants did not (Figure 4G, Supplemental Figure 7). Thus, RothiaRdnI bound and protected against ProteusRdnE, demonstrating that cross-protection between non-cognate EI pairs from different phyla is possible, provides a fitness benefit during competition, and influences community structure.

Given that the entire immunity protein is necessary for protection, we next evaluated which region(s) of RdnI allowed for cross-protection. We first moved the conserved motif of the three foreign RdnI homologs into ProteusRdnI-StrepII and measured neutralizing activity using swarm competition assays and binding activity using BACTH. The conserved motifs from Rothia and Prevotella were sufficient to preserve ProteusRdnI’s neutralizing (Figure 4H) and binding functions (Figure 4I). However, the conserved motif from Pseudomonas was insufficient to neutralize ProteusRdnE (Figure 4H), even though the construct could still bind ProteusRdnED39A (Figure 4I). We then moved the Proteus conserved motif into the RdnI variants from Prevotella and Pseudomonas, but the motif was insufficient to confer neutralizing (Figure 4H) or binding function (Figure 4I). These results showed that binding RdnE is necessary but not sufficient for RdnI protection. Further, within RdnI, the conserved motif is necessary for neutralization.

RdnE and RdnI proteins from diverse phyla are present in individual human microbiomes

Our findings indicate that immunity proteins such as RdnI could provide a broader protective umbrella for a cell beyond inhibiting the effector proteins of their siblings. If so, one would expect to find evidence of RdnE and RdnI homologs from different phyla in the same environment or microbial community. We tested this hypothesis by analyzing about 500,000 publicly available microbiomes (metagenomes) for the specific rdnE and rdnI genes examined in Figure 4 (Figure 5A). 2,296 human and terrestrial metagenomes contained reads matching with over 90% identity to these rdnE sequences (Figure 5B). The reads mapped to the expected niche for each organism, underscoring the presence of these specific effector proteins in naturally occurring human-associated microbiomes. Further, the rdnE and rdnI genes from various human-associated bacteria occurred concurrently in individual human oral and, to a lesser extent, gut metagenomes. The rdnE and rdnI genes from Rothia and Prevotella co-occurred in approximately 5% of the metagenomes analyzed (Figure 5C). Stringent detection parameters were utilized, so the true number could be higher. We then compared the abundance of rdnI to rdnE, since metagenomic coverage (i.e., the number of short reads that map to a gene) approximates the underlying gene’s abundance in the sampled community. In most gut samples, rdnI recruited more reads than rdnE, although there was substantial variance (Figure 5D). These data could indicate the presence of orphan rdnI genes, which is consistent with published T6SS orphan immunity alleles35, 36. These metagenomic patterns suggest that a single community can produce multiple RdnE and RdnI proteins from different phyla to defend against each other, further highlighting the intricate nature of bacterial defense in crowded environments.

The RdnI protein family has the potential for broader protection within oral and gut microbiomes.

A) Methodology used to identify rdnE and rdnI genes in publicly available metagenomic data. Metagenomes were mapped against sequences with a stringency of 90%. “Coverage” denotes the average depth of short reads mapping to a gene in a single sample. Colors represent rdnE and rdnI from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), or P. ogarae (dark blue). B) The experimentally tested rdnE gene sequences from different organisms (colors) are found in thousands of human-associated metagenomes. Each dot represents a single sample’s coverage of an individual rdnE gene, note log10-transormed y-axis. Only samples with >1x coverage are shown. C) Euler diagram showing the number of samples with co-occurring rdnE genes from different taxa (colors). D) Kernel density plot of the ratio of rdnI to rdnE coverage. The ratio of rdnI to rdnE was defined as log10(I/E) where I and E are the mean nucleotide’s coverage for rdnI and rdnE, respectively. The distribution of ratios was summarized as a probability density function (PDF) for each taxon (color) in each environment (subpanel). Here, the y-axis (unitless) reflects the probability of observing a given ratio (x-axis) in that dataset. The colored numbers in the top right of each panel show the number of metagenomes above the detection limit for both rdnE and rdnI for each taxon. Dashed vertical lines represent the median ratio. E) Skeleton-key model for immunity protein protection. Top, the current prevailing model for immunity proteins is that protection is defined by necessary and sufficient binding between cognate effectors (locks) and immunity proteins (keys). Bottom, our proposed skeleton-key model for protection is that multiple immunity proteins (skeleton-keys) can bind a single effector due to a flexible (promiscuous) binding site. Protection is a two-step process of binding and then neutralization.

Discussion

Using these results as a foundation, we propose an alternate model of “EI skeleton keys,” in which flexible (“promiscuous”) binding between immunity and effector proteins enables protection in mixed-species communities (Figure 5E). This model expands the current paradigm of (selective) cognate EI partners37 and has far-reaching implications for immunity protein evolution and molecular function. Unlike most known immunity proteins that fit within the prevailing model, we demonstrated that flexible EI binding is necessary but not sufficient to neutralize RdnE. We also showed that both RdnE and RdnI proteins have multiple domains, one conserved with functional activity and the other variable across phyla. The combined findings point to a two-step mechanism for RdnI protection: the N-terminal variable-sequence domain mediates binding to an effector, while the C-terminal conserved domain is required for neutralization in a function that has yet to be determined (Figure 5E).

The RdnE and RdnI protein families have parallel structural organizations. RdnE’s conserved region contains the enzymatic PD-(D/E)XK motif, which is necessary for DNA nuclease activity. Its variable region comprises the second half of the protein, which is also necessary for degradation activity. Despite amino acid variability among variants, these domains can be swapped between phyla and retain function. On the other hand, RdnI’s variable and conserved regions likely determine binding and function, respectively. ProteusRdnI’s first 85 amino acids are sufficient and necessary for binding ProteusRdnE, while the conserved motif, found in the last two-thirds of RdnI, is necessary for function. RdnI’s protective ability can be preserved when the conserved motif is swapped between species, even among those that cannot otherwise bind. The interchangeability between domains in RdnE and RdnI variants could reflect constraints on their evolutionary paths.

Compared to the prevailing model, our proposed skeleton-key model better explains RdnI’s conserved motif, orphan immunity genes, and other non-cognate protein interactions. Sequence diversity among EI pairs is often attributed to the coevolution between effector and immunity proteins, which is thought to maintain specificity and binding33. However, this coevolution model does not fully explain immunity proteins’ conserved regions or the presence and protective activity of orphan immunity proteins38, 39. A recent study with the Tri1 immunity protein shows that it has conserved enzymatic activity, allowing for protection against foreign effectors, but there is no cross-binding between non-cognate EI pairs40. This data suggests that conserved enzymatic activity can be maintained alongside the coevolution necessary for strict cognate pair binding. Therefore, we propose that EI protein domains may evolve independently instead of coevolution across the full protein. For example, RdnE’s PD-(D/E)XK motif and RdnI’s conserved motif might need to be maintained for activity, while the variable domains may diversify in sequence more freely. Depending on the selective pressures, the variable regions could reinforce specificity between cognate EI pairs as they coevolve, like the Tri1 family. Additional evolutionary analysis would reveal how the balance between specificity and flexibility evolves in EI pairs, both within domains and across the entire protein.

Flexible protection likely applies beyond contact-dependent EI pairs. Indeed, this model is conceptually similar to the cross-binding seen in type II antitoxins and bacteriocins. Within the ParD antitoxin family, 45% of the twenty antitoxins tested show cross-protective activity with at least one non-cognate ParE toxin41. For the E group of colicins, non-cognate immunity proteins show a lower binding affinity than for their cognate immunity proteins, but they still offer cross-protection13, 42. Similarly, immunity proteins from Carnobacterium piscicola and Enterococcus faecium offer cross-protection against the other’s cognate bacteriocin43. In these systems, as with RdnE-RdnI, flexible binding between a harmful protein and a non-cognate neutralizing protein can result in a potential fitness advantage for the bacteria.

When considering the impacts on community structure, the broadened activity of RdnI proteins against RdnE effectors from multiple phyla likely increases bacterial fitness, which is advantageous in dense environments. RdnI production increased cells’ fitness and modified the community in experimental setups like the swarm competition assays as it enabled vulnerable bacteria to inhabit previously restricted spaces. Supporting this experimental data, both gut and oral metagenomes showed evidence of multiple rdnE-rdnI pairs within individual samples, particularly between Rothia and Prevotella. Interestingly, the oral microbiome had roughly equivalent abundance between the effector and immunity genes, which might reflect that bacteria occupy distinct spatiotemporal niches within oral microbiomes, e.g. R. dentocariosa is predominantly on tooth surfaces44. By contrast, rdnI genes had greater abundance compared to rdnE in the gut microbiomes, which may reflect the greater diversity in member species and community structures found in the gut45. Orphan immunity genes are indeed a known phenomenon in T6SS EI literature but are usually documented through single isolate sequencing. This community level assessment affirms the presence of rdnI orphan genes on a population scale and points to relatively widespread immunity genes in hundreds or thousands of samples.

Given the ability of immunity genes to protect against non-cognate effectors, the presence of diverse orphan rdnI genes hints at the ecological complexity surrounding RdnE and RdnI. This community of immunity proteins is reminiscent of the model for shared immunity proteins within an ecosystem, also called a “hyper-immunity state,” which was seen among colicins in wild field mice46. In this hyper-immunity state, a set of immunity proteins shared among a community could offer an advantage against pathogens. Invading bacteria would be unable to protect themselves from certain effectors while the community would be protected as they share the immunity proteins necessary for defense. Flexible binding like RdnI could contribute to such a “hyper-immunity state” to help a bacterial community maintain its niche.

Indeed, bacteria have a diverse set of protective measures to ward off foreign effectors in addition to flexible immunity proteins. Recent work has identified non-specific mechanisms of protection including stress-response, physical barriers, and a stronger offense11. Orphan immunity genes also exist throughout many bacterial genomes and may be a part of this system38, for example, these genes offer a fitness advantage in mouse microbiomes39. Our data extends the current repertoire of protection mechanisms by adding another tool: a flexible immunity protein collection, where each immunity protein acts as a skeleton key against a wider class of effectors. This collection could be useful in dense, diverse communities where contact-dependent competition using EI pairs is critical to maintain one’s population. Taken together, the physical interactions between, and evolution of, effector and immunity proteins remain a rich area for new explorations.

Supporting information

eLife Assessment

This work presents valuable information about the specificity and promiscuity of toxic effector and immunity protein pairs. The evidence supporting the claims of the authors is currently incomplete, as there is concern about the methodology used to analyze protein interactions, which did not take potential differences in expression levels, protein folding, and/or transient interaction into account. Other methods to measure the strength of interactions and structural predictions would improve the study. The work will be of interest to microbiologists and biochemists working with toxin-antitoxin and effector-immunity proteins.

Reviewer #1 (Public Review):

In this manuscript, Knecht, Sirias et al describe toxin-immunity pair from Proteus mirabilis. Their observations suggest that the immunity protein could protect against non-cognate effectors from the same family. They analyze these proteins by dissecting them into domains and constructing chimeras which leads them to the conclusion that the immunity can be promiscuous and that the binding of immunity is insufficient for protective activity.

Strengths:

The manuscript is well written and the data are very well presented and could be potentially interesting. The phylogenetic analysis is well done, and provides some general insights.

Weaknesses:

1. Conclusions are mostly supported by harsh deletions and double hybrid assays. The later assays might show binding, but this method is not resolutive enough to report the binding strength. Proteins could still bind, but the binding might be weaker, transient, and out-competed by the target binding.

2. While the authors have modeled the structure of toxin and immunity, the toxin-immunity complex model is missing. Such a model allows alternative, more realistic interpretation of the presented data. Firstly, the immunity protein is predicted to bind contributing to the surface all over the sequence, except the last two alpha helices (very high confidence model, iPTM>0.8). The N terminus described by the authors contributes one of the toxin-binding surfaces, but this is not the sole binding site. Most importantly, other parts of the immunity protein are predicted to interact closer to the active site (D-E-K residues). Thus, based on the AlphaFold model, the predicted mechanism of immunization remains physically blocking the active site. However, removing the N terminal part, which contributes large interaction surface will directly impact the binding strength. Hence, the toxin-immunity co-folding model suggests that proper binding of immunity, contributed by different parts of the protein, is required to stabilize the toxin-immunity complex and to achieve complete neutralization. Alternative mechanisms of neutralization might not be necessary in this case and are difficult to imagine for a DNAse.

3. Dissection of a toxin into two domains is also not justified from a structural point of view, it is probably based on initial sequence analyses. The N terminus (actually previously reported as Pone domain in ref 21) is actually not a separate domain, but an integral part of the protein that is encased from both sides by the C terminal part. These parts might indeed evolve faster since they are located further from the active site and the central core of the protein. I am happy to see that the chimeric toxins are active, but regarding the conservation and neutralization, I am not surprised, that the central core of the protein fold is highly conserved. However, "deletion 2" is quite irrelevant - it deletes the central core of the protein, which is simply too drastic to draw any conclusions from such a construct - it will not fold into anything similar to an original protein, if it will fold properly at all.

4. Regarding the "promiscuity" there is always a limit to how similar proteins are, hence when cross-neutralization is claimed authors should always provide sequence similarities. This similarity could also be further compared in terms of the predicted interaction surface between toxin and immunity.

Overall, it looks more like a regular toxin-immunity couple, where some cross-reactions with homologues will, of course, be possible, depending on how far the sequences have deviated. Nevertheless, taking all of the above into account, these results do not challenge toxin-immunity specificity dogma.

Reviewer #2 (Public Review):

Summary:

The manuscript by Knecht et al entitled "Non-cognate immunity proteins provide broader defenses against interbacterial effectors in microbial communities" aims at characterizing a new type VI secretion system (T6SS) effector immunity pair using genetic and biochemical studies primarily focused on Proteus mirabilis and metagenomic analysis of human-derived data focused on Rothia and Prevotella sequences. The authors provide evidence that RdnE and RdnI of Proteus constitute an E-I pair and that the effector likely degrades nucleic acids. Further, they provide evidence that expression of non-cognate immunity derived from diverse species can provide protection against RdnE intoxication. Overall, this general line of investigation is underdeveloped in the T6SS field and conceptually appropriate for a broad audience journal. The paper is well-written and, aside from a few cases, well-cited. As detailed below however, there are several aspects of this paper where the evidence provided is somewhat insufficient to support the claims. Further, there are now at least two examples in the literature of non-cognate immunity providing protection against intoxication, one of which is not cited here (Bosch et al PMID 37345922 - the other being Ting et al 2018). In general therefore I think that the motivating concept here in this paper of overturning the predominant model of interbacterial effector-immunity cognate interactions is oversold and should be dialed back.

Strengths:

One of the major strengths of this paper is the combination of diverse techniques including competition assays, biochemistry, and metagenomics surveys. The metagenomic analysis in particular has great potential for understanding T6SS biology in natural communities. Finally, it is clear that much new biology remains to be discovered in the realm of T6SS effectors and immunity.

Weaknesses:

The authors have not formally shown that RdnE is delivered by the T6SS. Is it the case that there are not available genetics tools for gene deletion for the BB2000 strain? If there are genetic tools available, standard assays to demonstrate T6SS-dependency would be to interrogate function via inactivation of the T6SS (e.g. by deleting tssC).

For swarm cross-phyla competition assays (Figure 4), at what level compared to cognate immunity are the non-cognate immunity proteins being expressed? This is unclear from the methods and Figure 4 legend and should be elaborated upon. Presumably these non-cognate immunity proteins are being overexpressed. Expression level and effector-to-immunity protein stoichiometry likely matters for interpretation of function, both in vitro as well as in relevant settings in nature. It is important to assess if native expression levels of non-cognate cross-phyla immunity (e.g. Rothia and Prevotella) protect similarly as the endogenously produced cognate immunity. This experiment could be performed in several ways, for example by deleting the RdnE-I pair and complementing back the Rothia or Prevotella RdnI at the same chromosomal locus, then performing the swarm assay. Alternatively, if there are inducible expression systems available for Proteus, examination of protection under varying levels of immunity induction could be an alternate way to address this question. Western blot analysis comparing cognate to non-cognate immunity protein levels expressed in Proteus could also be important. If the authors were interested in deriving physical binding constants between E and various cognate and non-cognate I (e.g. through isothermal titration calorimetry) that would be a strong set of data to support the claims made. The co-IP data presented in supplemental Figure 6 are nice but are from E. coli cells overexpressing each protein and do not fully address the question of in vivo (in Proteus) native expression.

Lines 321-324, the authors infer differences between E and I in terms of read recruitment (greater abundance of I) to indicate the presence of orphan immunity genes in metagenomic samples (Figure 5A-D). It seems equally or perhaps more likely that there is substantial sequence divergence in E compared to the reference sequence. In fact, metagenomes analyzed were required only to have "half of the bases on reference E-I sequence receiving coverage". Variation in coverage again could reflect divergent sequence dipping below 90% identity cutoff. I recommend performing metagenomic assemblies on these samples to assess and curate the E-I sequences present in each sample and then recalculating coverage based on the exact inferred sequences from each sample.

A description of gene-level read recruitment in the methods section relating to metagenomic analysis is lacking and should be provided.

Reviewer #3 (Public Review):

Summary:

The authors discovered that the RdnE effector possesses DNase activity, and in competition, P. mirabilis having RdnE outcompetes the null strain. Additionally, they presented evidence that the RdnI immunity protein binds to RdnE, suppressing its toxicity. Interestingly, the authors demonstrated that the RdnI homolog from a different phylum (i.e., Actinomycetota) provides cross-species protection against RdnE injected from P. mirabilis, despite the limited identity between the immunity sequences. Finally, using metagenomic data from human-associated microbiomes, the authors provided bioinformatic evidence that the rdnE/rdnI gene pair is widespread and present in individual microbiomes. Overall, the discovery of broad protection by non-cognate immunity is intriguing, although not necessarily surprising in retrospect, considering the prolonged period during which Earth was a microbial battlefield/paradise.

Strengths:

The authors presented a strong rationale in the manuscript and characterized the molecular mechanism of the RdnE effector both in vitro and in the heterologous expression model. The utilization of the bacterial two-hybrid system, along with the competition assays, to study the protective action of RdnI immunity is informative. Furthermore, the authors conducted bioinformatic analyses throughout the manuscript, examining the primary sequence, predicted structural, and metagenomic levels, which significantly underscore the significance and importance of the EI pair.

Weaknesses:

1. The interaction between RdnI and RdnE appears to be complex and requires further investigation. The manuscript's data does not conclusively explain how RdnI provides a "promiscuous" immunity function, particularly concerning the RdnI mutant/chimera derivatives. The lack of protection observed in these cases might be attributed to other factors, such as a decrease in protein expression levels or misfolding of the proteins. Additionally, the transient nature of the binding interaction could be insufficient to offer effective defenses.

2. The results from the mixed population competition lack quantitative analysis. The swarm competition assays only yield binary outcomes (Yes or No), limiting the ability to obtain more detailed insights from the data.

3. The discovery of cross-species protection is solely evident in the heterologous expression-competition model. It remains uncertain whether this is an isolated occurrence or a common characteristic of RdnI immunity proteins across various scenarios. Further investigations are necessary to determine the generality of this behavior.

Comments from Reviewing Editor:

In addition to the references provided by Reviewer #2, the first manuscript to show non-cognate binding of immunity proteins was Russell et al 2012 (PMID: 22607806).

IdrD was shown to form a subfamily of effectors in this manuscript by Hespanhol et al 2022 PMID: 36226828 that analyzed several T6SS effectors belonging to PDDExK, and it should be cited.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment

This work presents valuable information about the specificity and promiscuity of toxic effector and immunity protein pairs. The evidence supporting the claims of the authors is currently incomplete, as there is concern about the methodology used to analyze protein interactions, which did not take potential differences in expression levels, protein folding, and/or transient interaction into account. Other methods to measure the strength of interactions and structural predictions would improve the study. The work will be of interest to microbiologists and biochemists working with toxin-antitoxin and effector-immunity proteins.

We thank the reviewers for considering this manuscript. We agree that this manuscript provides a valuable and cross-discipline introduction to new EI pair protein families where we focus on the EI pair’s flexibility and impacts on community structure. As such, we believe we have provided a solid foundation for future studies to examine non-cognate interactions and their possible effects on microbial communities. This, by definition, leaves some areas “incomplete” and, therefore, open for further investigations. While the methods we show do consider potential differences in binding assays, we have more explicitly addressed how “expression, protein folding, and/or transient binding” may play into this expanded EI pair model. We have also tempered the discussion of the proposed model, while also clearly highlighting other published evidence of non-cognate binding interactions between effector and immunity proteins. We have responded to the reviewers’ public comments (italicized below).

In this revised manuscript, we have updated the main text, particularly the Discussion section, to include more careful language, explain past research better, and add new references to works showing non-cognate immunity proteins protecting against effectors in other systems. We have also updated the supplemental files with more analyses; the relevant procedures are in the Materials and Methods.

Public Reviews:

Note: Reviewer 1, who appeared to focus on a subset of the manuscript rather than the whole, based their comments on several inaccuracies, which we discuss below. We found the tone in this reviewer's comments to be, at times, inappropriate, e.g., using "harsh" and "simply too drastic" to imply that common structure-function analyses were outside of the field-standard methods. We also note that the reviewer took a somewhat atypical step in reviewing this manuscript by running and analyzing the potential protein-complex data in AlphaFold2 but did not discuss areas of low confidence within that model that may contradict their conclusions. We are concerned their approach muddled valid scientific criticisms with problematic conclusions.

Reviewer #1 (Public Review):

In this manuscript, Knecht, Sirias et al describe toxin-immunity pair from Proteus mirabilis. Their observations suggest that the immunity protein could protect against non-cognate effectors from the same family. They analyze these proteins by dissecting them into domains and constructing chimeras which leads them to the conclusion that the immunity can be promiscuous and that the binding of immunity is insufficient for protective activity.

Strengths:

The manuscript is well written and the data are very well presented and could be potentially interesting. The phylogenetic analysis is well done, and provides some general insights.

Weaknesses:

(1) Conclusions are mostly supported by harsh deletions and double hybrid assays. The later assays might show binding, but this method is not resolutive enough to report the binding strength. Proteins could still bind, but the binding might be weaker, transient, and out-competed by the target binding.

The phrasing of structure-function analyses as “harsh” is a bit unusual, as other research groups regularly use deletions and hybrid studies. Given the known caveats to deletion and domain substitutions, we included point-mutation analyses for both the effector and immunity proteins, as found on lines 105 - 113 and 255 - 261 in the current manuscript. These caveats are also why we coupled the in vitro binding analyses with in vivo protection experiments in two distinct experimental systems (E. coli and P. mirabilis). Based on this manuscript’s introductory analysis (where we define and characterize the genes, proteins, interactions, phylogenetics, and incidences in human microbiomes), the next apparent questions are beyond the scope of this study. Future approaches would include analyzing purified proteins from the effector (E) and immunity (I) protein families using biochemical assays, such as X-ray crystallography, circular dichroism spectroscopy, among others.

Interestingly, most papers in the EI field do not measure EI protein affinity (Jana et al., 2019, Yadav et al., 2021). Notable exceptions are earlier colicin research (Wallis et al., 1995) and a new T6SS EI paper (Bosch et al., 2023) published as we first submitted this manuscript.

(2) While the authors have modeled the structure of toxin and immunity, the toxin-immunity complex model is missing. Such a model allows alternative, more realistic interpretation of the presented data. Firstly, the immunity protein is predicted to bind contributing to the surface all over the sequence, except the last two alpha helices (very high confidence model, iPTM>0.8). The N terminus described by the authors contributes one of the toxin-binding surfaces, but this is not the sole binding site. Most importantly, other parts of the immunity protein are predicted to interact closer to the active site (D-E-K residues). Thus, based on the AlphaFold model, the predicted mechanism of immunization remains physically blocking the active site. However, removing the N terminal part, which contributes large interaction surface will directly impact the binding strength. Hence, the toxin-immunity co-folding model suggests that proper binding of immunity, contributed by different parts of the protein, is required to stabilize the toxin-immunity complex and to achieve complete neutralization. Alternative mechanisms of neutralization might not be necessary in this case and are difficult to imagine for a DNase.

In response to the reviewer’s comment, we again reviewed the RdnE-RdnI AlphaFold2 complex predictions with the most updated version of ColabFold (1.5.2-patch with PDB100 and MMseq2) and have included them at the end of these responses [1].

However, the literature reports that computational predictions of E-I complexes often do not match experimental structural results (Hespanhol et al., 2022, Bosch et al., 2023). As such, we chose not to include the predicted cognate and non-cognate RdnE-I complexes from ColabFold (which uses AlphaFold2) and have not included this data in the revised manuscript. (It is notable that reviewer 1 found the proposed expanded model and research so interesting as to directly input and examine the AI-predicted RdnE-RdnI protein interactions in AlphaFold2.)

Discussion of the prevailing toxin-immunity complex model is in the introduction (lines 45-48) and Figure 5E. Further, there are various known mechanisms for neutralizing nucleases and other T6SS effectors, which we briefly state in the discussion (lines 359 - 361). More in-depth, these molecular mechanisms include active-site blocking (Benz et al., 2012), allosteric-site binding (Kleanthous et al., 1999 and Lu et al., 2014), enzymatic neutralization of the target (Ting et al., 2021), and structural disruption of both the active and binding sites (Bosch et al., 2023). Given this diversity of mechanisms, we did not presume to speculate on the as-of-yet unknown mechanism of RdnI protection. We have expanded discussion of these items in the revised manuscript.

(3) Dissection of a toxin into two domains is also not justified from a structural point of view, it is probably based on initial sequence analyses. The N terminus (actually previously reported as Pone domain in ref 21) is actually not a separate domain, but an integral part of the protein that is encased from both sides by the C terminal part. These parts might indeed evolve faster since they are located further from the active site and the central core of the protein. I am happy to see that the chimeric toxins are active, but regarding the conservation and neutralization, I am not surprised, that the central core of the protein fold is highly conserved. However, "deletion 2" is quite irrelevant - it deletes the central core of the protein, which is simply too drastic to draw any conclusions from such a construct - it will not fold into anything similar to an original protein, if it will fold properly at all.

The reviewer’s comment highlights why we turned to the chimera proteins to dissect the regions of RdnE (formerly IdrD-CT), as the deletions could result in misfolded proteins. (We initially examined RdnE in the years before the launch of AlphaFold2.) However, the reviewer is incorrect regarding the N-terminus of RdnE. The PoNe domain, while also a subfamily of the PD-(D/E)XK superfamily, forms a distinct clade of effectors from the PD-(D/E)XK domain in RdnE (formally IdrD-CT) as seen in Hespanhol et al., 2022; this is true for other DNase effectors as well. Many studies analyzing effectors within the PD-(D/E)XK superfamily only focus on the PD-(D/E)XK domain, removing just this domain from the context of the whole protein (Hespanhol et al., 2022; Jana et al., 2019). Of note, in RdnE, this region alone (containing the DNA-binding domain) is insufficient for DNase activity (unlike in PoNe). We have clarified this distinction in the results section of the current manuscript, visible in figure 2 .

(4) Regarding the "promiscuity" there is always a limit to how similar proteins are, hence when cross-neutralization is claimed authors should always provide sequence similarities. This similarity could also be further compared in terms of the predicted interaction surface between toxin and immunity.

Reviewer 1 points out a fundamental property of protein-protein interactions that has been isolated away from the impacts of such interactions on bacterial community structure. We have provided the whole protein alignments in figure 3 supplemental figure 3, the summary images in Figure 3D, and the protein phylogenetic trees in Figure 3C. We encourage others to consider the protein alignments as percent amino acid sequence similarity is not necessarily a good gauge for protein function and interactions. These data are publicly available on the OSF website associated with this manuscript https://osf.io/scb7z/, and we hope the community explores the data there.

In consideration of the enthusiasm to deeply dive into the primary research data, we have included the pairwise sequence identities across the entire proteins here: Proteus RdnI vs. Rothia RdnI: 23.6%; Proteus RdnI vs. Prevotella RdnI: 16.3%, Proteus RdnI vs. Pseudomonas RdnI: 14.6%; Rothia RdnI vs. Prevotella RdnI: 22.4%, Rothia RdnI vs. Pseudomonas RdnI: 17.6%; Prevotella RdnI vs. Pseudomonas RdnI: 19.5%. (As stated in response to reviewer 1 comment 2, we did not find it appropriate to make inferences based on AlphaFold2-predicted protein complexes.)

Overall, it looks more like a regular toxin-immunity couple, where some cross-reactions with homologues are possible, depending on how far the sequences have deviated. Nevertheless, taking all of the above into account, these results do not challenge toxin-immunity specificity dogma.

In this manuscript, we did not intend to dismiss the E-I specificity model but rather point out its limitations and propose an important expansion of that model that accounts for cross-protection and survival against attacks from other genera. We agree that it is commonly considered that deviations in amino acid sequence over time could result in cross-binding and protection (see lines 364-368). However, the impacts of such cross-binding on community structure, bacterial survival, and strain evolution were rarely addressed in prior literature, with exceptions such as in Zhang et al., 2013 and Bosch et al., 2023 among others. One key insight we propose and show in this manuscript is that cross-binding can be a fitness benefit in mixed communities; therefore, it could be selected for evolutionarily (lines 378-380), even potentially in host microbiomes.

Reviewer #2 (Public Review):

Summary:

The manuscript by Knecht et al entitled "Non-cognate immunity proteins provide broader defenses against interbacterial effectors in microbial communities" aims at characterizing a new type VI secretion system (T6SS) effector immunity pair using genetic and biochemical studies primarily focused on Proteus mirabilis and metagenomic analysis of human-derived data focused on Rothia and Prevotella sequences. The authors provide evidence that RdnE and RdnI of Proteus constitute an E-I pair and that the effector likely degrades nucleic acids. Further, they provide evidence that expression of non-cognate immunity derived from diverse species can provide protection against RdnE intoxication. Overall, this general line of investigation is underdeveloped in the T6SS field and conceptually appropriate for a broad audience journal. The paper is well-written and, aside from a few cases, well-cited. As detailed below however, there are several aspects of this paper where the evidence provided is somewhat insufficient to support the claims. Further, there are now at least two examples in the literature of non-cognate immunity providing protection against intoxication, one of which is not cited here (Bosch et al PMID 37345922 - the other being Ting et al 2018). In general therefore I think that the motivating concept here in this paper of overturning the predominant model of interbacterial effector-immunity cognate interactions is oversold and should be dialed back.

We agree that analyses focusing on flexible non-cognate interactions and protection are underdeveloped within the T6SS field and are not fully explored within a community structure. These ideas are rapidly growing in the field, as evidenced by the references provided by the reviewer. As stated earlier, we did not intend to overturn the prevailing model but rather have proposed an expanded model that accounts for protection against attacks from foreign genera.

Strengths:

One of the major strengths of this paper is the combination of diverse techniques including competition assays, biochemistry, and metagenomics surveys. The metagenomic analysis in particular has great potential for understanding T6SS biology in natural communities. Finally, it is clear that much new biology remains to be discovered in the realm of T6SS effectors and immunity.

Weaknesses:

The authors have not formally shown that RdnE is delivered by the T6SS. Is it the case that there are not available genetics tools for gene deletion for the BB2000 strain? If there are genetic tools available, standard assays to demonstrate T6SS-dependency would be to interrogate function via inactivation of the T6SS (e.g. by deleting tssC).

Our research group showed that the T6SS secretes RdnE (previously IdrD) in Wenren et al., 2013 (cited in lines 71-73). We later confirmed T6SS-dependent secretion by LC-MS/MS (Saak et al., 2017).

For swarm cross-phyla competition assays (Figure 4), at what level compared to cognate immunity are the non-cognate immunity proteins being expressed? This is unclear from the methods and Figure 4 legend and should be elaborated upon. Presumably these non-cognate immunity proteins are being overexpressed. Expression level and effector-to-immunity protein stoichiometry likely matters for interpretation of function, both in vitro as well as in relevant settings in nature. It is important to assess if native expression levels of non-cognate cross-phyla immunity (e.g. Rothia and Prevotella) protect similarly as the endogenously produced cognate immunity. This experiment could be performed in several ways, for example by deleting the RdnE-I pair and complementing back the Rothia or Prevotella RdnI at the same chromosomal locus, then performing the swarm assay. Alternatively, if there are inducible expression systems available for Proteus, examination of protection under varying levels of immunity induction could be an alternate way to address this question. Western blot analysis comparing cognate to non-cognate immunity protein levels expressed in Proteus could also be important. If the authors were interested in deriving physical binding constants between E and various cognate and non-cognate I (e.g. through isothermal titration calorimetry) that would be a strong set of data to support the claims made. The co-IP data presented in supplemental Figure 6 are nice but are from E. coli cells overexpressing each protein and do not fully address the question of in vivo (in Proteus) native expression.

P. mirabilis strain ATCC29906 does not encode the rdnE and rdnI genes on the chromosome (NCBI BioSample: SAMN00001486) (line 151). Production of the RdnI proteins, including the cognate Proteus RdnI, comes from equivalent transgenic expression vectors. Specifically, the rdnI genes were expressed under the flaA promoter in P. mirabilis strain ATCC29906 (Table 1) for the swarm competition assays found in Figure 2C and Figure 4. This promoter results in constitutive expression in swarming cells (Belas et al., 1991; Jansen et al., 2003). In the revised manuscript, figure 4 Supplement Figure 2 shows the relative RdnI protein levels in these strains; we also clarified the expression constructs in the text (see reviewer 3, comment 1).

Lines 321-324, the authors infer differences between E and I in terms of read recruitment (greater abundance of I) to indicate the presence of orphan immunity genes in metagenomic samples (Figure 5A-D). It seems equally or perhaps more likely that there is substantial sequence divergence in E compared to the reference sequence. In fact, metagenomes analyzed were required only to have "half of the bases on reference E-I sequence receiving coverage". Variation in coverage again could reflect divergent sequence dipping below 90% identity cutoff. I recommend performing metagenomic assemblies on these samples to assess and curate the E-I sequences present in each sample and then recalculating coverage based on the exact inferred sequences from each sample.

This comment raises the challenges with metagenomic analyses. It was difficult to balance specificity to a particular species’ DNA sequence with the prevalence of any homologous sequence in the sample. Given the distinction in binding interactions among the examined four species, we opted to prioritize specificity, accepting that we were losing access to some rdnE and rdnI sequences in that decision. We chose a 90% identity cutoff, which, through several in silica controls, ensured that each sequence we identified was the rdnE or rdnI gene from that specific species. For the Version of Record, we have included analysis with a 70% cutoff in the supplemental information to try to account for sequence divergence by lowering the identity cutoffs as suggested. The data from the 70% identity cutoff was consistent with the original data from the 90% identity cutoff.

A description of gene-level read recruitment in the methods section relating to metagenomic analysis is lacking and should be provided.

Noted. We included the raw code and sequences on the OSF website associated with this manuscript https://osf.io/scb7z/.

Reviewer #3 (Public Review):

Summary:

The authors discovered that the RdnE effector possesses DNase activity, and in competition, P. mirabilis having RdnE outcompetes the null strain. Additionally, they presented evidence that the RdnI immunity protein binds to RdnE, suppressing its toxicity. Interestingly, the authors demonstrated that the RdnI homolog from a different phylum (i.e., Actinomycetota) provides cross-species protection against RdnE injected from P. mirabilis, despite the limited identity between the immunity sequences. Finally, using metagenomic data from human-associated microbiomes, the authors provided bioinformatic evidence that the rdnE/rdnI gene pair is widespread and present in individual microbiomes. Overall, the discovery of broad protection by non-cognate immunity is intriguing, although not necessarily surprising in retrospect, considering the prolonged period during which Earth was a microbial battlefield/paradise.

Strengths:

The authors presented a strong rationale in the manuscript and characterized the molecular mechanism of the RdnE effector both in vitro and in the heterologous expression model. The utilization of the bacterial two-hybrid system, along with the competition assays, to study the protective action of RdnI immunity is informative. Furthermore, the authors conducted bioinformatic analyses throughout the manuscript, examining the primary sequence, predicted structural, and metagenomic levels, which significantly underscore the significance and importance of the EI pair.

Weaknesses:

(1) The interaction between RdnI and RdnE appears to be complex and requires further investigation. The manuscript's data does not conclusively explain how RdnI provides a "promiscuous" immunity function, particularly concerning the RdnI mutant/chimera derivatives. The lack of protection observed in these cases might be attributed to other factors, such as a decrease in protein expression levels or misfolding of the proteins. Additionally, the transient nature of the binding interaction could be insufficient to offer effective defenses.

Yes, we agree with the reviewer and hope that grant reviewers’ share this colleague’s enthusiasm for understanding the detailed molecular mechanisms of RdnE-RdnI binding across genera. In the revised manuscript, we have continued to emphasize such caveats as the next frontier is clearly understanding the molecular mechanisms for RdnI cognate or non-cognate protection. In the revised manuscript, figure 4 Supplement Figure 2 shows the RdnI protein levels; we also clarified the expression constructs in the text (see reviewer 2, comment 2).

(2) The results from the mixed population competition lack quantitative analysis. The swarm competition assays only yield binary outcomes (Yes or No), limiting the ability to obtain more detailed insights from the data.

The mixed swam assay is needed when studying T6SS effectors that are primarily secreted during Proteus’ swarming activity (Saak et al. 2017, Zepeda-Rivera et al. 2018). This limitation is one reason we utilize in vitro, in vivo, and bioinformatic analyses. Though the swarm competition assay yields a binary outcome, we are confident that the observed RdnI protection is due to interaction with a trans-cell RdnE via an active T6SS. By contrast, many manuscripts report co-expression of the EI pair (Yadev et al., 2021, Hespanhol et al., 2022) rather than secreted effectors, as we have achieved in this manuscript.

(3) The discovery of cross-species protection is solely evident in the heterologous expression-competition model. It remains uncertain whether this is an isolated occurrence or a common characteristic of RdnI immunity proteins across various scenarios. Further investigations are necessary to determine the generality of this behavior.

We agree, which is why we submitted this paper as a launching point for further investigations into the generality of non-cognate interactions and their potential impact on community structure.

Comments from Reviewing Editor:

In addition to the references provided by reviewer#2, the first manuscript to show non-cognate binding of immunity proteins was Russell et al 2012 (PMID: 22607806).

IdrD was shown to form a subfamily of effectors in this manuscript by Hespanhol et al 2022 PMID: 36226828 that analyzed several T6SS effectors belonging to PDDExK, and it should be cited.

We appreciate that the reviewer and eLife staff pointed out missed citations. We have incorporated these studies and cited them in the revised manuscript.

[1] The Proteus RdnE in complex with either the Prevotella or Pseudomonas RdnI showed low confidence at the interface (pIDDT ~50-70%); this AI-predicted complex might support the lack of binding seen in the bacterial two-hybrid assay. On the other hand, the Proteus and Rothia RdnI N-terminal regions show higher confidence at the interface with RdnE. Despite this, the C-terminus of the Proteus RdnI shows especially low confidence (pIDDT ~50%) where it might interact near RdnE’s active site (as suggested by reviewer 1). Given this low confidence and the already stated inaccuracies of AI-generated complexes, we would rather wait for crystallization data to inform potential protection mechanisms of RdnI.

Version 2

Abstract

Abstract

Dense microbial communities, like the gut and soil microbiomes, are dynamic societies. Bacteria can navigate these environments by deploying proteins that alter foreign cells’ behavior, such as interbacterial effectors. Current models suggest that adjacent sibling cells are protected by an immunity protein, as compared to toxin-antitoxin systems that act only within the effector-producing cell. A prevailing hypothesis is that immunity proteins binding to specific (cognate) protein partners is sufficient to disrupt effector function. Further, there is little-to-no crosstalk with other non-cognate effectors. In this research, we build on sporadic reports challenging these hypotheses. We show that immunity proteins from a newly defined protein family can bind and protect against non-cognate PD-(D/E)XK-containing effectors from diverse phyla. We describe the domains essential for binding and function and show that binding alone is insufficient for protective activity. Moreover, we found that these effector and immunity genes co-occur in individual human microbiomes. These results expand the growing repertoire of bacterial protection mechanisms and the models on how non-cognate interactions impact community structure within complex ecosystems.

Introduction

Specificity between protein-protein interactions is key for many biological processes, such as metabolism, development, and intercellular signaling. Binding to an incorrect partner or disrupted binding of the specific (cognate) protein can cause a diseased state or cell death1. For bacterial social behaviors, cognate protein-protein interactions between cells impact organism fitness and community structure, such as excluding foreign cells2. This importance of specificity between cognate partners in social behaviors remains largely unexplored. However, in other contexts, flexible (“promiscuous”) binding allows protein partners to retain their interactions when undergoing rapid mutational changes, such as during immune recognition of viral particles3,4. Unknown is whether flexible binding between noncognate proteins can occur during bacterial social behaviors and thereby impact microbial communities. An expanded protective function would reveal new bacterial behaviors that influence individual fitness and community structure in dynamic ecosystems.

Microbes often exist within dense, multi-phyla communities, like the human gut microbiome, where they communicate and compete with neighbors. Bacteria can use effector-immunity protein (EI) pairs in these environments to gain advantages5,6. Unlike bacterial toxin-antitoxin (TA) systems in which a single cell produces both toxic and neutralizing proteins, bacteria inject cell-modifying proteins (called interbacterial “effectors”) directly into neighboring cells via several contact-dependent transport mechanisms, including the type VI secretion system (T6SS), type IV secretion system (T4SS), and contact-dependent inhibition (CDI)7,8. Clonal siblings produce the matching immunity protein that modifies the effector’s activity. For lethal effectors, both clonal and non-clonal cells are negatively impacted when binding is disrupted or absent5. These interactions between EI pairs can shape community composition by changing bacterial fitness.

The interaction specificity between matching EI pairs has historically defined immunity protein protection, but recent studies raise doubts. Of note, EI pairs interact within a neighboring cell which creates unique restrictions for both their protection mechanisms and their evolution9. Currently, the predominant model is that T6SS-associated EI pairs act like a tumbler lock-and-key, where each effector protein has a single cognate partner10. Immunity proteins bind their cognate effectors, often at the active site, to neutralize effector activity11,12. However, experiments with engineered proteins reveal that small amino acid sequence changes to an immunity protein can allow it to bind effectors other than its cognate partner13,14. Also, the T6SS-associated effector and immunity proteins from Salmonella enterica subsp. enterica serovar Typhimurium and Enterobacter cloacae, which are phylogenetically close, bind each other in vitro and protect against the other in vivo15. Another example is Tde1 and Tdi1. Homologous Tdi1 immunity proteins lacking a cognate effector (i.e., “orphans”) bound and protected against the effector from a different organism16. These studies indicate that the widely used tumbler lock-and-key model does not account for the potential breadth of immunity protein protection.

We studied an EI pair in Proteus mirabilis to examine this prevailing model. This opportunistic pathogen resides in human and animal guts and can cause recurrent and persistent urinary tract infections17. P. mirabilis encodes two T6SS-dependent EI pairs (one lethal and one non-lethal) that impact collective motility and relative fitness18,19. For the lethal EI pair, previously termed Idr19, the molecular functions remained unknown. Here, we characterized this EI pair and determined the critical residues for activity, leading to the identification of two protein families. We showed that proteins in the immunity protein family bind non-cognate effectors produced by bacteria from different phyla and result in altered population structures. Structure-function assays revealed that a conserved region within the C-terminus of the immunity proteins is necessary to neutralize the P. mirabilis effector protein. Further, we found that the flexible EI pairs from various phyla naturally co-occur in individual human microbiomes. These findings provide compelling evidence for cross-protection and support a critical revision of the model for EI pairs, particularly in consideration of ecological significance.

Results

RdnE is a DNA nuclease and seeds a PD-(D/E)XK subfamily

To compete against other strains, P. mirabilis strain BB2000 requires both the idrD gene and the T6SS, suggesting that the idrD-encoded protein functions as a T6SS-associated effector18. The idrD gene contains an Rhs region within its N-terminus. Many Rhs-containing effectors often contain an enzymatic domain in the C-terminus20,21. As a result, we investigated the function of the final 138 amino acids at IdrD’s C-terminus, now renamed “RdnE” for recognition DNA nuclease effector. We measured bacterial growth using a strain derived from BB2000 that has disruptions in its native idrD and downstream genes19. This P. mirabilis culture had 1000 fewer cells per mL when engineered to overproduce RdnE in trans than the negative control containing the parent empty vector (Figure 1A). An equivalent growth pattern occurred in Escherichia coli cells under the same conditions (Figure 1 - Figure supplement 1). Thus, RdnE was lethal in vivo.

RdnE homologs act as DNA endonucleases and contain interchangeable domains.

A) Cell viability (colony forming units [CFU] per mL) after protein production in swarms of P. mirabilis strain idrD*, which does not produce RdnE and RdnI. Cells produced GFPmut2, RdnE, or mutant variants in the predicted PD-(D/E)XK motif: D39A, E53A, K55A, or all. B) In vitro DNA degradation assay for ProteusRdnE. Increasing concentrations of a negative control, ProteusRdnE-FLAG, or ProteusRdnED39A-FLAG were incubated with methylated or unmethylated lambda DNA (48,502 bp) and analyzed by gel electrophoresis. Plasmid DNA degradation is in Figure 1 - Figure supplement 1. C) In vitro DNA degradation assay for domain deletions of ProteusRdnE. The first construct removed the first alpha helix without disturbing the catalytic residues, and the second construct contained the PD-(D/E)XK motif and removed region 2. Increasing concentrations were analyzed as in (B). D) Multiple sequence alignment between P. mirabilis and R. dentocariosa RdnE sequences. The black bar marks the PD-(D/E)XK motif, and the gray bar marks the variable region 2 domain. Conserved residues are highlighted in dark blue. Secondary structure predictions identified using Ali2D55,56 (h for alpha helix, e for beta sheet); the catalytic residues (stars) are noted above the alignment. (E,F) In vitro DNA degradation assay and analysis as in (B). (E) Increasing concentrations of either a negative control, RothiaRdnE-FLAG, or RothiaRdnED39A-FLAG. (F) The PD-(D/E)XK motifs were swapped between the RothiaRdnE (orange bar) and the ProteusRdnE (green bar) sequences and compared to the wild-type RdnE proteins.

RdnE, an endonuclease, is lethal in Escherichia coli and cuts plasmid DNA.

A) Growth curve of E. coli cells overexpressing ProteusRdnE or variants with mutations in the PD-(D/E)K active site. Cells were grown at 37°C for 16 hours. Optical density at 595 nm (OD595) was measured every half hour. A control strain expressing an empty vector was used as the negative control. B) Micrographs of E. coli cells producing ProteusRdnE-FLAG or ProteusRdnED39A-FLAG, isolated during mid-logarithmic growth and imaged. DAPI was used to detect DNA within cells. Top, the empty vector as a negative control. Middle, E. coli producing ProteusRdnE from an inducible plasmid. Bottom, E. coli producing ProteusRdnED39A from an inducible plasmid. Left, phase; right, fluorescence. C) Anti-FLAG Western Blot for ProteusRdnE-FLAG and ProteusRdnED39A-FLAG generated by in vitro translation. Protein levels were determined by comparison to a standard dilution of FLAG-BAP. A negative control (DHFR) without a FLAG tag was also produced with the in vitro translation reaction. A vertical orange line separates the membrane where the ladder was marked with a pencil after transfer (to the left) and the membrane after western blot detection (to the right). D) In vitro DNase assay reactions on cut and uncut plasmid DNA. In vitro translation products of either a negative control (DHFR), ProteusRdnE-FLAG, or ProteusRdnED39A-FLAG were incubated with cut or uncut plasmid DNA and analyzed with gel electrophoresis.

RdnE’s initial 86 amino acids contain a PD-(D/E)XK motif, which is suggestive of nucleotide degradation. The PD-(D/E)XK superfamily includes proteins with broad functions, including effectors that degrade DNA or RNA22–24. Three residues in the catalytic site—D, D/E, and K—are required for activity25. Therefore, we changed the corresponding residues in RdnE (D39, E53, and K55) to alanine, separately and together. P. mirabilis producing these mutant proteins showed growth equivalent to the negative control lacking RdnE (Figure 1A). We also saw that E. coli cells that were producing RdnE had morphologies that were indicative of DNA damage or stress, consistent with an SOS response26,27. These cells were elongated, and the DAPI-stained DNA was distributed irregularly within the cells (Figure 1 - Figure supplement 1). Cells producing a D39A mutant (RdnED39A) largely did not have this appearance, although a few elongated cells remained, suggesting that the D39A mutant retained partial activity (Figure 1 - Figure supplement 1). Therefore, the PD-(D/E)XK motif was essential for cell death.

The importance of the PD-(D/E)XK motif for activity suggested that RdnE was a nuclease, but defining its molecular target required direct analysis. Due to its lethality in P. mirabilis and E. coli cells, we synthesized RdnE with a C-terminus FLAG epitope tag using in vitro translation, which resulted in nanogram quantities (Figure 1 - Figure supplement 1). We added phage lambda DNA (methylated or unmethylated) to progressively higher RdnE protein concentrations and then performed agarose gel electrophoresis analysis. Degradation of lambda DNA occurred in the presence of RdnE, regardless of the DNA methylation state (Figure 1B). The RdnED39A construct caused a slight reduction in lambda DNA, while the negative control showed no DNA loss (Figure 1B). RdnE also caused a reduction in plasmid DNA, indicating it has endonuclease activity (Figure 1 - Figure supplement 1). These results revealed that RdnE caused DNA degradation in vitro in a PD-(D/E)XK-dependent manner.

RdnE appeared to have two different domains, as a region directly follows the PD-(D/E)XK motif. A two-domain architecture is similar to that described for DNases28,29. Yet, the PD-(D/E)XK domain could also be sufficient for DNase activity of some effectors such as PoNe-containing DNases22,30. Therefore, we examined whether RdnE’s PD-(D/E)XK motif was sufficient for DNA degradation or whether both domains were required for activity. We made independent deletions of each potential RdnE domain. One construct deleted the first alpha helix without disturbing the catalytic residues; the other deleted the region after the PD-(D/E)XK motif, which we termed “region 2.” The resulting proteins, produced via in vitro translation, were assayed for DNase activity as described above. The truncated proteins resulted in no loss of lambda DNA (Figure 1C), indicating that both domains were necessary for degradation activity.

We next asked whether RdnE homologs also act as DNA nucleases. A bioinformatics search revealed the closest RdnE homolog outside of Proteus was found in the Actinobacteria, Rothia dentocariosa C6B. Rothia species are inhabitants of the normal oral flora, dwelling in biofilms within the human oral cavity and pharynx31. The two RdnE proteins (ProteusRdnE and RothiaRdnE) share approximately 55% amino acid sequence identity, mostly within the PD-(D/E)XK domain; they have similar predicted secondary structures (Figure 1D). Given this, we hypothesized that RothiaRdnE also acted as a DNA nuclease.

We analyzed RothiaRdnE for PD-(D/E)XK-dependent DNA nuclease activity by producing it and a predicted null mutant, RothiaRdnED39A, using in vitro translation. Samples containing the RothiaRdnED39A protein or a negative control had similar DNA levels (Figure 1E). By contrast, samples with the wild-type RothiaRdnE protein showed a loss of lambda DNA regardless of methylation state, indicating that RothiaRdnE also had DNA nuclease activity (Figure 1E). Given that region 2 was necessary for activity in ProteusRdnE but has greater amino acid sequence diversity than the PD-(D/E)XK domain between the two proteins, we queried whether domains from foreign organisms could complement one another. We exchanged region 2 between the ProteusRdnE and RothiaRdnE sequences and assayed for nuclease activity. The hybrid proteins degraded lambda DNA, unlike the negative control (Figure 1F), demonstrating the cross-phyla protein domains could complement one another. Altogether, these findings demonstrated that these RdnE proteins form a PD-(D/E)XK-containing DNA nuclease subfamily. This conclusion is also consistent with recent literature showing that RdnE-containing proteins (formerly IdrD-CT32) form their own sub-clade within other PD-(D/E)XK-containing nucleases30.

RdnI binds and neutralizes RdnE

As effectors have cognate immunity proteins that are often located adjacently on the chromosome, we hypothesized that rdnI (formerly “idrE”), which is the gene directly downstream of rdnE in P. mirabilis (Figure 2A), encodes the cognate immunity protein. RdnI did not have defined domains, and its function was unknown. We assessed RdnI’s activity using microscopic and cell growth analysis. Swarming P. mirabilis cells are normally elongated with DAPI-stained DNA found along the cell body (Figure 2B). By contrast, swarming cells producing RdnE in trans did not elongate, had a reduced DAPI signal, and had an accumulation of misshapen cells (Figure 2B). Cell shape and DNA-associated fluorescence levels returned to normal when cells concurrently produced the RdnE and RdnI proteins (Figure 2B). RdnI production also rescued cell growth in E. coli cells producing RdnE (Figure 2 - Figure supplement 1). These data suggested that RdnI inhibits RdnE’s lethality.

RdnI binds to and protects against RdnE in vivo and in vitro.

A) Domain architecture for the idr locus in P. mirabilis strain BB200018. At the top are genes with Pfam domains listed below them. Gray boxes denote PAAR and Rhs domains in the N-terminal region of the full-length IdrD protein. B) Micrographs of P. mirabilis strain idrD* cells carrying an empty vector, a vector for producing RdnE, or a vector for producing RdnE and RdnI. DNA was visualized by DAPI stain. Phase, left; fluorescence, right. C) Swarm competition assay18 of wild-type P. mirabilis strain BB2000 (donor) competed against the vulnerable target, which is P. mirabilis strain ATCC29906 carrying an empty vector, a vector for producing RdnI-StrepII, or a vector for producing GFP, both under the fla promoter. Left: schematic of swarm competition assay where top left colony is BB2000, top right colony is ATCC29906 with its vector cargo, and bottom colony is a 1:1 mixture of BB2000 and ATCC29906 with its vector cargo. Gray boxes underneath indicate whether BB2000 (top) or ATCC29906 (bottom) dominate in the 1:1 mixture and white arrows point to a boundary line that forms between different strains. D) Bacterial two-hybrid (BACTH) assay with RdnED39A-FLAG, RdnI-StrepII, and GFPmut2. The colorimetric change was discerned in the presence of the substrate X-gal and inducer IPTG. E) An anti-FLAG batch co-immunoprecipitation of RdnED39A-FLAG and RdnI-StrepII. RdnED39A-FLAG or exogenous FLAG-BAP (soluble fraction) was incubated with anti-FLAG resin (FLAG flow through). RdnI-StrepII was then added to the resin (RdnI-StrepII flow through). Any proteins bound to resin were eluted with FLAG-peptide (Elution) and analyzed by anti-FLAG and anti-StrepII western blots.

RdnI offers protection against and binds to RdnE.

A) Viability assays of E. coli cells after production of RdnE, RdnI, or co-production of RdnE and RdnI within a cell. Cells were assayed for colony forming units per milliliter over a six-hour time course. B) The Coomassie blue-stained gel for the anti-FLAG batch co-immunoprecipitation assay results shown in the main text, Figure 2E.

We next evaluated whether RdnI provided protection against injected RdnE within mixed communities similar to native ecosystems. We used well-established swarm competition assays, which combine one-to-one mixtures of P. mirabilis strains to measure dominance in two-dimensional population structures19. The control strain was wild-type strain BB2000 (herein called “BB2000”), which naturally produces RdnE and RdnI. The other was strain ATCC29906, which does not naturally produce RdnE and RdnI. These two strains formed a visible boundary between swarming monoculture colonies (Figure 2C). The mixed-strain colony merged with BB2000 in one-to-one competitions, demonstrating BB2000’s dominance in the two-dimensional population structure (Figure 2C). A similar outcome was seen when ATCC29906 produced a vector-encoded Green Fluorescent Protein (GFPmut2) under the fla promoter, which results in constitutive gene expression in swarming P. mirabilis cells33,34. However, BB2000 did not outcompete ATCC29906 engineered to produce vector-encoded RdnI with a C-terminal Strep-tag II epitope tag (“RdnI-StrepII”) under the fla promoter; this is visible in the merging of the mixed-strain colony with ATCC29906 (Figure 2C). Thus, RdnI protected cells against injected RdnE within mixed communities.

Based on the prevailing EI model, we predicted that a cognate EI pair should bind to one another, which we evaluated in vivo and in vitro. We used the attenuated mutant (RdnED39A-FLAG) for these assays because producing the wild-type RdnE protein kills cells. For in vivo analysis, we used bacterial two-hybrid assays (BACTH) in which the reconstitution of the T18 and T25 fragments of adenylate cyclase results in the colorimetric change to blue in the presence of the substrate, X-gal35,36. Constructed vectors contained genes for RdnED39A-FLAG, RdnI-StrepII, or GFPmut2 on the C-termini of the T18 or the T25 fragment. When the reporter strain produced RdnED39A-FLAG or RdnI-StrepII with GFPmut2, the resultant yellow color was equivalent to when X-gal was absent (Figure 2D). There was also minimal color change when an individual protein was produced on both fragments (Figure 2D). However, the reporter strains made blue colonies when X-gal was present, and the cells concurrently produced RdnED39A-FLAG and RdnI-StrepII. These results indicated that RdnE and RdnI bind to each other in vivo.

We used batch in vitro co-immunoprecipitation assays to confirm the in vivo binding result. Separate E. coli strains produced either RdnED39A-FLAG or had a negative control, exogenous FLAG-BAP (E. coli bacterial alkaline phosphatase with a FLAG epitope tag) added to cell lysate. An anti-FLAG western blot showed both FLAG-BAP (∼50 kDa) and RdnED39A-FLAG (∼17 kDa) in the soluble and elution fractions. RdnI-StrepII eluted with RdnED39A-FLAG but not the negative control (Figure 2E). The western blot results corresponded with the Coomassie blue-stained gels (Figure 2 - Figure supplement 1). Overall, our data showed that Proteus RdnE and RdnI form a cognate EI pair with impacts on population structure. Questions about their prevalence among bacteria and their ecological relevance remained.

Expansion of the RdnE and RdnI protein families revealed similar gene architecture and secondary structures.

Gene neighborhood analysis can guide homology inference and protein comparisons. We conducted consecutive searches with BLAST37 and HMMER38 to identify sequences that encoded proteins with high similarity to RdnE and RdnI (Figure 3 - Figure supplement 1). The final list contained 21 EI pairs from a variety of phyla that are located adjacently in their respective genomes (Table 1). Although the genes surrounding these putative EI pairs differed, many shared mobile-associated elements, such as Rhs sequences or other similar peptide-repeat sequences (Figure 3A). Several gene neighborhoods had secretion-associated genes, such as the T6SS-associated vgrG/tssI gene and the CDI-associated cdiB gene. A few also included putative immunity proteins from other reported families, like immunity protein 44 (Pfam15571) in Taylorella asinigenitalis MCE3 and immunity protein 51 (Pfam15595) in Chryseobacterium populi CF314. Notably, these organisms varied widely in origin and residence. Some were from the soil rhizosphere (Pseudomonas ogarae and C. populi) and others from the human microbiome (P. mirabilis, R. dentocariosa, and Prevotella jejuni) (Figure 3A). The prevalence of these genes across the phylogenetic tree (Figure 3B) and the presence of secretion-associated loci in the gene neighborhoods suggested a role in cell-cell interactions and potentially community structure.

RdnE and RdnI protein families share conserved residues and predicted structures.

A) Gene neighborhoods for RdnE and RdnI homologs. Listed are gene neighborhoods, relevance, and niche, which we identified using IMG/M from the Joint Genomics Institute. Colors highlight conserved function/genes (not to scale). (Agr: Agriculture, Med: Medical, Env: Environmental), and the site of isolation. B) Phylogenetic tree based on NCBI taxonomy. Scale is located below the graph. The colored circles represent phyla (green: Actinobacteriota; yellow: Firmicutes; blue: Bacteroidota; pink: Proteobacteria). C) Unrooted maximum likelihood trees of the RdnE (left) and RdnI (right) homologs. Trees were created with RaxML70, and the scale is annotated below. The colored circles represent phyla (same as in B). D) Protein alignments overlaid with either predicted secondary structures (top) or conserved residues (bottom) of the RdnE and RdnI homologs. MUSCLE alignments63 are highlighted by secondary structures (red: alpha helices, light blue: beta sheets), or conserved residues (dark blue). White represents gaps in the protein alignment. The bars below mark the predicted conserved and variable domains. E) Alignments of AlphaFold2 predictions for RdnE and RdnI sequences from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), and P. ogarae (dark blue). Structures were generated using ColabFold40 and aligned using PyMol.

RdnE and RdnI protein families show conserved structures.

A) Diagram detailing the methodology used for identifying sequences homologous to RdnE and RdnI. Seven homologs of ProteusRdnE (orange) found using BLAST37 with their corresponding downstream genes (pink) were aligned and used as seed for a sequential search using HMMER search38 and the Ensembl database64. Gene neighborhoods were analyzed for genomes with adjacent rdnE and rdnI genes. B) Tanglegram71 of the RdnE and RdnI protein families from the 21 sequences. On the left is the maximum-likelihood tree for the RdnE protein family on the right is the maximum-likelihood tree for the RdnI protein family. Black lines match effector and immunity pairs from the same species. C-F) MUSCLE alignment63 of RdnE (C and D) and RdnI (E and F) protein families highlighted with either predicted secondary structure predictions (C and E) or conserved residues (D and F). Secondary structures were predicted using Ali2D65,66 and are shaded by confidence. Predicted α-helices in pink; β-strands in light blue. Conserved residues were highlighted (dark blue) using Jalview63. Black lines underneath mark the truncated variants of RdnI described in the main text Figure 4.

AlphaFold2 predictions for RdnE and RdnI homologs.

A) Confidence scores (pIDDT) for AlphaFold239,40 predictions for RdnE sequences from P. mirabilis, R. dentocariosa, P. jejuni, and P. ogarae. The confidence score (y-axis) for each residue (x-axis) are graphed for the five ranked models (rank 1: blue, rank 2: orange, rank 3: green, rank 4: red, rank 5: purple). B) RdnE AlphaFold2 rank 1 models. Models were colored by confidence scores. Red indicates high confidence (90-100%) while blue indicates low confidence (30-50%). C) Confidence scores for AlphaFold2 predictions of RdnI homologs from P. mirabilis, R. dentocariosa, P. jejuni, and P. ogarae. D) We colored the AlphaFold2 rank 1 model, including both the BB2000 sequence and natural variant, by confidence scores where red is high confidence (90-100%) and blue is low confidence (30-50%).

Given the diversity in species, we next examined the relationship between the various RdnE- and RdnI-like proteins and whether there was syntony between these proteins given that they are encoded adjacently on each identified genome. We constructed maximum likelihood trees to examine the relationship between the identified RdnE- and RdnI-like proteins. The RdnE and RdnI trees diverged from the species tree (Figure 3B). However, overall, the arrangement of RdnE- and RdnI-like proteins within the maximum likelihood trees was similar and showed syntony (Figure 3 - Figure supplement 1), though small differences were present (Figure 3C). For example, P. mirabilis and R. dentocariosa proteins shared more similarities than to those from more closely related genera. These results are consistent with the potential horizontal gene transfer reported for other EI pairs7.

Given these results, we reasoned that the domain architectures and amino acid diversity could reveal functions for the two families. When we examined the predicted secondary structures of the RdnE-like proteins, they were conserved despite differences in the amino acid sequences (Figure 3D, Figure 3 - Figure supplement 1). The RdnE proteins showed two distinct domains, as with the Proteus and Rothia results (Figure 1): the PD-(D/E)XK region followed by a sequence variable region (region 2). Further, the AlphaFold2-generated39,40 predicted tertiary structures were consistent with PD-(D/E)XK folds (three β-sheets flanked by two α-helices, α/β/α) found in other proteins (Figure 3E, Figure 3 - Figure supplement 2)25. These findings suggested that domains in RdnE are conserved across diverse phyla and further confirm that the sequences seed a distinct PD-(D/E)XK subfamily.

While immunity proteins within a family have diverse overall amino acid sequences, conserved secondary structures and some conserved residues are common in some immunity protein families. Indeed, they are often used to characterize these families41. We found that while the RdnI proteins shared minimal primary amino acid sequence identity, they were predicted to contain several α-helices (Figure 3D, Figure 3 - Figure supplement 1) and had similar AlphaFold2-predicted tertiary structures (Figure 3E, Figure 3 - Figure supplement 2). We also discovered a region with three alpha-helices and several conserved residues, which we named the “conserved motif” (Figure 3D). The RdnI conserved motif might be a key domain for seeding this novel immunity protein family.

Binding flexibility in RdnI allows for cross-species protection

We deployed a structure-function approach to determine the conserved motif’s role in ProteusRdnI’s activity. Analysis using AlphaFold239,40 and Consurf42 revealed seven highly conserved residues within this region; four of these (Y197, H221, P244, E246) clustered together within the AlphaFold2 structure and are identical between sequences (Figures 4A). In a sequence-optimized (SO) RdnI, we independently changed each of these four residues to alanine and discovered that each alanine-substituted variant behaved like the wildtype and inhibited RdnE activity (Figure 4 - Figure supplement 1). We then replaced all seven residues (Y197, S235, K258, and the original four) with alanine (ProteusRdnI7mut-StrepII) and found that, unlike wild-type ProteusRdnI-StrepII, this construct was not protective in swarm competition assays (Figure 4B). However, the ProteusRdnI7mut-StrepII mutant still bound RdnED39A-FLAG in bacterial two-hybrid assays (Figure 4C). Therefore, the seven residues in the conserved motif are critical for RdnI’s neutralizing function but dispensable for binding RdnE.

The RdnI protein family can offer cross-protection due to an interchangeable conserved domain that is critical for function.

A) Sequence logo of the RdnI’s conserved motif. Stars indicate the seven analyzed residues. B) Swarm competition assay with ATCC29906 producing either RdnI-StrepII or RdnI7mut-StrepII, which contains mutations in all seven conserved residues. We used a sequence-optimized (SO) RdnI protein that had a higher GC% content and an identical amino acid sequence for ease of cloning. Left: schematic of swarm competition assay as in Figure 2. Gray boxes indicate which strain dominated over the other. White arrows point to the boundary formed between different strains. C) BACTH assay of RdnED39A-FLAG with SO RdnI-StrepII or RdnI7mut-StrepII. GFPmut2 was used as a negative control. D) Swarm competition assay with ATCC29906 expressing either the wild-type RdnI or a RdnI truncation. The three truncations were in the first alpha helix (amino acids 1-85), the second half of RdnI (amino acids 150-305), and the end of the protein (amino acids 235-305). E) BACTH assay of RdnED39A-FLAG with wild-type RdnI and the three RdnI truncations. F) Swarm competition assay with ATCC29906 expressing foreign RdnI proteins. G) BACTH assay of RdnED39A-FLAG with each of the foreign RdnI proteins. GFPmut2 was used as a negative control. H) Swarm competition assay with ATCC29906 producing SO RdnI with swapped conserved motifs. I) BACTH assay of RdnED39A-FLAG with SO RdnI with swapped conserved motifs. Colored bars denote RdnI-StrepII proteins from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), or P. ogarae (dark blue).

Single mutations in the RdnI conserved motif do not alter protective function.

Swarm competition assay19 with single residue mutations in the conserved motif of RdnI. P. mirabilis BB2000 (donor) was competed against the vulnerable P. mirabilis ATCC29906 expressing RdnI with single residue mutations in four of the seven conserved residues (highlighted in Figure 4A) in ProteusRdnI-StrepII (ProteusRdnIY197A-StrepII, ProteusRdnIH221A-StrepII, ProteusRdnIY244A-StrepII, ProteusRdnIY246A-StrepII). All constructs were made in a sequence-optimized RdnI.

RdnI protein levels are similar under constitutive fla promoter in P. mirabilis.

Swarm cell protein expression assay2 on P. mirabilis ATCC29906 cells expressing each of the four RdnI proteins under the constitutive fla promoter. Soluble fraction and whole cell extract samples were then run on SDS-Page gels and incubated with anti-StrepII antibodies (left) or Coomassie blue (right). 20 ng of GFP-StrepII (Iba Lifesciences, Gӧttingen Germany) was used as a positive control.

Anti-FLAG co-IPs reveal mixed binding results for foreign immunity protein

Anti-FLAG co-immunoprecipitation assay between ProteusRdnED39A-FLAG and the RdnI-StrepII proteins from P. mirabilis, R. dentocariosa, P. jejuni, or P. ogarae. RdnED39A-FLAG was incubated with anti-FLAG resin (FLAG soluble fraction). RdnI-StrepII containing lysate was then added to the resin (RdnI-StrepII flow through). Proteins bound to resin were then eluted with FLAG-peptide (elution). The negative control was exogenous FLAG-BAP. Samples were incubated with either anti-FLAG antibodies (A), anti-StrepII antibodies (B), or were stained with Coomassie blue (C).

Given that the conserved motif and nearby regions are likely involved in protective activity, we queried for potential functions in the remainder of the RdnI protein. We engineered variants that were either (1) the first 85 amino acids, (2) amino acids 150 to 305, which contained an intact conserved motif, or (3) amino acids 235 to 305, which contained the last alpha helix of the conserved motif (Figure 3 - Figure supplement 1). None of these constructs protected against RdnE’s lethality in vivo during the swarm competition assay (Figure 4D), demonstrating that the entire protein is likely essential for function. However, the variant containing the first 85 amino acids of ProteusRdnI was the only construct to bind ProteusRdnE, indicating that the N-terminal region is sufficient for binding between this P. mirabilis EI pair (Figure 4E). Thus, our data suggests that binding is necessary but not sufficient for neutralization. Also, the inhibitory activity might reside within the second half of RdnI. As the prevailing model defines cognate-specificity by binding activity, our structure-function results for RdnE (Figure 1F) and RdnI (Figure 4E) suggest that this model does not fully explain the complex interactions between effectors and immunity proteins.

Therefore, we explored the relationship between non-cognate RdnE and RdnI proteins from various phyla. We first asked whether non-cognate RdnI immunity proteins could protect against injected ProteusRdnE. Using the swarm competition assays, we competed BB2000 against ATCC29906 engineered to produce vector-encoded RdnI homologs from P. mirabilis, R. dentocariosa, P. jejuni, or P. ogarae (ProteusRdnI-StrepII, RothiaRdnI-StrepII, PrevotellaRdnI-StrepII, and PseudomonasRdnI-StrepII, respectively) under the fla promoter (Figure 2). BB2000 dominated the swarm when ATCC29906 produced GFPmut2, PrevotellaRdnI-StrepII, or PseudomonasRdnI-StrepII (Figure 4F). However, ATCC29906 outcompeted BB2000 when making ProteusRdnI-StrepII or RothiaRdnI-StrepII (Figure 4F). Expression levels of the transgenic RdnI proteins in ATCC29906 were similar (Figure 4 - Figure supplement 2). Further, the RdnI immunity proteins from Proteus and Rothia consistently bound ProteusRdnED39A in the in vivo and in vitro assays; the Prevotella and Pseudomonas variants did not (Figure 4G, Figure 4 - Figure supplement 3). The binding to and protection of RothiaRdnI against ProteusRdnE demonstrated that cross-protection between non-cognate EI pairs from different phyla is possible, provides a fitness benefit during competition, and influences community structure.

While overall the EI pairs showed syntony with each other (Figure 3 - Figure supplement 1), amino acid changes can have critical impacts on whether there are specific or flexible interactions between non-cognate protein pairs4. Therefore, we next evaluated which region(s) of RdnI contributes to cross-protection. We first moved the conserved motif of the three foreign RdnI homologs into ProteusRdnI-StrepII and measured neutralizing activity using swarm competition assays and binding activity using BACTH. The conserved motifs from Rothia and Prevotella were sufficient to preserve ProteusRdnI’s neutralizing (Figure 4H) and binding functions (Figure 4I). However, the conserved motif from Pseudomonas was not sufficient to neutralize ProteusRdnE (Figure 4H), even though the construct still bound ProteusRdnED39A (Figure 4I). We then moved the Proteus conserved motif into the RdnI variants from Prevotella and Pseudomonas. These Prevotella and Pseudomonas hybrid proteins did not protect against ProteusRdnE in the swarm competition assay (Figure 4H) or bind to it in the BACTH assay (Figure 4I), indicating that the conserved motif is not required for binding but, alone, is insufficient to confer protection. Thus, RdnI-like immunity proteins containing this conserved motif can protect against non-cognate effector proteins if binding has been established.

RdnE and RdnI proteins from diverse phyla are present in individual human microbiomes

Our findings revealed that immunity proteins such as RdnI could provide a broader protective umbrella for a cell beyond inhibiting the effector proteins of their siblings. If so, one would expect to find evidence of RdnE and RdnI homologs from different phyla in the same environment or microbial community. We tested this hypothesis by analyzing around 500,000 publicly available microbiomes (metagenomes) for the specific rdnE and rdnI gene sequences examined in Figure 4 (Figure 5A). 2,296 human and terrestrial metagenomes contained reads matching with over 90% identity to these rdnE sequences (Figure 5B). We used this cutoff to ensure that each nucleotide sequence queried in the metagenomes closely matched experimentally characterized reference sequences. As a control, we applied a 70% identity threshold, which would retain related but more divergent sequences. We saw similar patterns with a total ∼ 2% change in the number of genomes per category (Figure 5 - Figure supplement 1). The reads mapped to the expected niche for each organism, underscoring the presence of the genes encoding these specific effector proteins in naturally occurring human-associated microbiomes.

The RdnI protein family has the potential for broader protection within oral and gut microbiomes.

A) Methodology used to identify rdnE and rdnI genes in publicly available metagenomic data. Metagenomes were mapped against sequences with a stringency of 90%. “Coverage” denotes the average depth of short reads mapping to a gene in a single sample. Colors represent rdnE and rdnI from P. mirabilis (green), R. dentocariosa (orange), P. jejuni (magenta), or P. ogarae (dark blue). B) The experimentally tested rdnE gene sequences from different organisms (colors) are found in thousands of human-associated metagenomes. Each dot represents a single sample’s coverage of an individual rdnE gene, note log10-transormed y-axis. Only samples with >1x coverage are shown. C) Euler diagram showing the number of samples with co-occurring rdnE genes from different taxa (colors). D) Kernel density plot of the ratio of rdnI to rdnE coverage. The ratio of rdnI to rdnE was defined as log10(I/E) where I and E are the mean nucleotide’s coverage for rdnI and rdnE, respectively. The distribution of ratios was summarized as a probability density function (PDF) for each taxon (color) in each environment (subpanel). Here, the y-axis (unitless) reflects the probability of observing a given ratio (x-axis) in that dataset. The colored numbers in the top right of each panel show the number of metagenomes above the detection limit for both rdnE and rdnI for each taxon. Dashed vertical lines represent the median ratio. E) Skeleton-key model for immunity protein protection. Top, the current prevailing model for T6SS immunity proteins is that protection is defined by necessary and sufficient binding between cognate effectors (locks) and immunity proteins (keys). Bottom, a proposed, expanded model: multiple immunity proteins (skeleton-keys) can bind a single effector due to a flexible (promiscuous) binding site. Protection is a two-step process of binding and then neutralization.

Metagenomic analysis with a 70% stringency revealed similar patterns in RdnE and RdnI localization.

The same metagenomic analysis as described in Figure 5 was used but had a lower stringency (70% instead of 90% identity to the rdnE and rdnI sequences). A) Each dot represents a single sample’s coverage of an individual rdnE gene, note log10-transormed y-axis. Only samples with >1x coverage are shown. B) Euler diagram showing the number of samples with co-occurring rdnE genes from different taxa (colors). C) Kernel density plot of the ratio of rdnI to rdnE coverage. The ratio of rdnI to rdnE was defined as log10(I/E) where I and E are the mean nucleotide’s coverage for rdnI and rdnE, respectively. The distribution of ratios was summarized as a probability density function (PDF) for each taxon (color) in each environment (subpanel). Here, the y-axis (unitless) reflects the probability of observing a given ratio (x-axis) in that dataset. The colored numbers in the top right of each panel show the number of metagenomes above the detection limit for both rdnE and rdnI for each taxon. Dashed vertical lines represent the median ratio.

RdnE and RdnI sequences are found in metagenomic datasets.

A) Heatmap of the log10-normalized coverage of rdnE and rdnI (rows) from the focal taxa for all metagenomes (columns) where any were detected. Metagenomes are sorted by decreasing rdnE coverage. B) Span chart showing the difference in coverage between cognate rdnE (circles) to rdnI (crosses) for all taxa (colors) for metagenomes in which rdnE-rdnI from multiple taxa were detected.

The rdnE and rdnI genes from various human-associated bacteria occurred concurrently in individual human oral and, to a lesser extent, gut metagenomes. The rdnE and rdnI genes from Rothia and Prevotella co-occurred in approximately 5% of the metagenomes analyzed (Figure 5C, Figure 5 - Figure supplement 2). Stringent detection parameters were utilized, so the true number could be higher. We then compared the abundance of rdnI to rdnE reads, since metagenomic coverage (i.e., the number of short reads that map to a gene) approximates the underlying gene’s abundance in the sampled community. In most gut samples, rdnI recruited more reads than rdnE, although there was substantial variance (Figure 5D). These data could indicate the presence of orphan rdnI genes, which is consistent with published T6SS orphan immunity alleles16,43,44. These metagenomic patterns suggest that a single community can produce multiple RdnE and RdnI proteins from different phyla, thereby providing a potential for them to interact in a host environment.

Discussion

Using these results as a foundation, we propose an extension to the prevailing model of selective, cognate EI partners45 to incorporate “EI skeleton keys.” In this revised model, flexible (“promiscuous”) binding between non-cognate effector and immunity proteins enables broader protection in mixed-species communities (Figure 5E). Here, we demonstrated that RdnE-RdnI binding is necessary but not sufficient to neutralize RdnE, which differs from many previously described T6SS immunity proteins. We showed that full-length RdnE, containing both its PD-(D/E)XK domain and variable C-terminal region, is a DNA-degrading endonuclease. Likewise, RdnI requires its entire protein to bind and neutralize RdnE, including a newly identified conserved C-terminal motif. Our findings point to a possible two-step mechanism for how the RdnI immunity protein works: the N-terminal variable-sequence domain mediates binding to an effector, while the C-terminal conserved domain contributes to neutralization in a not-yet-determined mechanism (Figure 5E). These findings have potential impacts on our understanding of immunity protein evolution, molecular functions, and microbial community structure.

The domain architectures of RdnE and RdnI suggest possible evolutionary trajectories for these EI pairs. While both require the full-length protein, RdnE and RdnI can function with residue changes within the domains and even retain activity in cross-phyla hybrid proteins. Tri1 immunity proteins also contain two domains, but these domains are associated with distinct functions. Their conserved region corresponds with the broad-acting enzymatic domain, which is independent of their cognate-specific binding domain,46 suggesting that conserved enzymatic activity can be maintained alongside potential coevolution necessary for strict cognate pair binding. Therefore, individual domains in EI proteins may evolve independently rather than the entire protein experiencing coevolution. As such, RdnE’s PD-(D/E)XK motif and RdnI’s conserved motif might be maintained for activity, while the variable domains may diversify in sequence independently. Depending on the selective pressures, the variable regions could reinforce specificity between cognate EI pairs as they coevolve. Additional evolutionary analysis would reveal how the balance between specificity and flexibility evolves in EI pairs, both within domains and across the entire protein.

RdnI’s potential two-step mechanism adds to a growing number of ways in which immunity proteins neutralize effector proteins. However, RdnI’s neutralization mechanism remains unknown. Many crystalized structures of EI complexes show that immunity proteins can bind and occlude an effector’s active site, effectively neutralizing the effector’s function12,47. However, some immunity proteins allosterically inhibit their effector without blocking the active site48,49. Recent studies revealed more neutralization mechanisms. The Tri1 immunity protein has a conserved enzymatic function that neutralizes its effector’s activity, allowing for protection against foreign effectors in addition to selective cognate-binding activity,46 while the Tdi1 immunity protein conformationally disrupts its effector, Tde1’s, binding and active sites to prevent DNA nuclease activity16. One will need to experimentally determine the structure of the RdnE-RdnI complex is necessary to define how it neutralizes RdnI and how this molecular mechanism compares to other immunity proteins.

Regardless, our results indicate that RdnI’s conserved domain is essential for protective activity and that a combination of seven highly conserved residues mediates that protection. There are several possibilities for how this domain aids neutralization, such as an ion-binding pocket, structural stability, or protein partner binding. Another possibility is that RdnI’s conserved region reinforces binding to the effector, aiding non-cognate interactions or co-evolving pairs. For example, while our assays indicate the primary binding domain for RdnI is in the N-terminus, the conserved domain could reinforce an initial, transient binding interaction. Indeed, multiple binding domains have been recorded for TA systems and likely protect the cell during co-evolution50. Binding affinities between effector and immunity proteins are not well-documented; those reported vary. For DNA nuclease colicins, non-cognate interactions have affinities in the nanomolar range whereas cognate interactions have picomolar affinities51. However, the T6SS-associated EI pair, Tde1 and Tdi1, have similar nanomolar affinities for both cognate and non-cognate orphan EI pairs16. In addition, it is unclear what equilibrium dissociation constant (KD) for EI binding would confer protection in native systems. This KD may be especially important in the case of highly motile bacteria, such as P. mirabilis, that only need to survive long enough to escape. Additionally, interbacterial effectors act within a neighboring cell, which may make determining the native ratios of effector to immunity proteins challenging. It will be interesting to see how binding affinities between other EI pairs compare, both cognate and non-cognate interactions, and how these affinities relate to protection in mixed communities.

When considering the impacts on community structure, the broadened activity of RdnI proteins against RdnE effectors from multiple phyla likely increases bacterial fitness, which is advantageous in dense environments. Our analysis measured protection during a two-dimensional, swarm-structured competition, where RdnI offered a susceptible strain protection against trans-cellular RdnE delivered natively. As such, we can conclude that RdnI production increased individual cells’ fitness and modified the community structure; it enabled vulnerable bacteria to inhabit previously restricted spaces. Supporting this experimental data, both gut and oral metagenomes showed evidence of multiple rdnE-rdnI pairs within individual samples, particularly between Rothia and Prevotella. Interestingly, the oral microbiome had roughly equivalent abundance between the effector and immunity genes, which might reflect that bacteria occupy distinct spatiotemporal niches within oral microbiomes, e.g. R. dentocariosa is predominantly on tooth surfaces52. By contrast, rdnI genes had greater abundance compared to rdnE in the gut microbiomes, which may reflect the greater diversity in member species and community structures found in the gut53. Orphan immunity genes are indeed a known phenomenon in T6SS EI literature but are usually documented through single isolate sequencing, with notable exceptions such as the Ross et al. 2019 and Bosch et al. 2023 papers. This community-level assessment affirms the presence of rdnI orphan genes on a population scale and points to relatively widespread immunity genes in hundreds or thousands of samples.

Given the ability of immunity genes to protect against non-cognate effectors, the presence of diverse orphan rdnI genes hints at the ecological complexity surrounding RdnE and RdnI. This community of immunity proteins is reminiscent of the model for shared immunity proteins within an ecosystem, called a “hyper-immunity state,” which was seen among colicins in wild field mice54. In this hyper-immunity state, a set of immunity proteins shared among a community could offer an advantage against pathogens. Invading bacteria would be unable to defend themselves from certain effectors, while the community would be protected as they share the collective immunity proteins. Flexible binding like RdnI could contribute to such a “hyper-immunity state” to help a bacterial community maintain its niche.

Indeed, bacteria have a diverse set of protective measures to ward off foreign effectors in addition to flexible immunity proteins. Recent work has identified non-specific mechanisms of protection including stress-response, physical barriers, and a stronger offense10. Orphan immunity genes also exist throughout many bacterial genomes and may be a part of this system,55 for example, orphan immunity genes offer a fitness advantage in vitro12,16 and in mouse microbiomes56. Flexible EI pairs are also not limited to secretion systems but are also seen among TA pairs57 and bacteriocins51,58. Our data extends the current repertoire of protection mechanisms by adding another tool: a flexible immunity protein collection, where each immunity protein acts as a skeleton key against a wider class of effectors. This flexibility is seen among orphan immunity proteins12,16 and for immunity proteins with cognate effectors as in this study. Flexible binding could be a general property of T6SS immunity proteins that could be useful in dense, diverse communities, like human and soil microbiomes, where contact-dependent competition using EI pairs is critical to maintain one’s population. Indeed, the physical interactions between, and evolution of, effector and immunity proteins remain a rich area for new explorations.